---
title: "India's Trade Story: Exports, Imports, Deficits, and Sectoral Composition"
description: "A data-driven analytical article on India's merchandise trade trajectory — with interactive ggiraph charts and gt tables drawn from HS code classified trade data."
author: "Pawan"
date: "2025/10/29"
date-modified: last-modified
categories: ["Trade","Economy","Export Policy","Articles"]
format:
  html:
    toc: true
    toc-depth: 3
    theme: flatly
    code-fold: true
editor_options:
  chunk_output_type: console
execute:
  error: false
  message: false
  warning: false
  eval: true
---

```{r setup, include=FALSE}
library(data.table); library(ggplot2); library(ggiraph)
library(readxl); library(viridis); library(scales); library(gt)

my_theme <- theme_minimal(base_size=13)+
  theme(text=element_text(color="#2c3e50"),
        plot.title=element_text(size=15,face="bold",hjust=0.5),
        plot.subtitle=element_text(size=11,hjust=0.5,color="#7f8c8d"),
        plot.caption=element_text(size=9,color="#95a5a6",hjust=0),
        axis.title=element_text(size=11,face="bold"),
        axis.text=element_text(size=10),
        axis.text.x=element_text(angle=45,vjust=1,hjust=1),
        legend.position="bottom",legend.title=element_blank(),
        panel.grid.major=element_line(color="#ecf0f1",linewidth=0.4),
        panel.grid.minor=element_blank(),
        strip.background=element_rect(fill="#f8f9fa",color=NA),
        strip.text=element_text(face="bold",size=10))

std_opts <- list(
  opts_tooltip(css=paste0("background-color:#2c3e50;color:white;padding:8px 12px;",
               "border-radius:6px;font-size:12px;line-height:1.5;")),
  opts_hover(css="fill-opacity:0.85;stroke:white;stroke-width:1.5;"),
  opts_hover_inv(css="opacity:0.25;"),
  opts_zoom(min=0.5,max=3), opts_toolbar(position="topright"),
  opts_sizing(rescale=TRUE,width=1))

usd <- as.data.table(readRDS("/home/pawan/trade/totaltrade/total_exportsusd.rds"))
qty <- as.data.table(readRDS("/home/pawan/trade/totaltrade/total_exportsqty.rds"))
musd <- as.data.table(readRDS("/home/pawan/trade/totaltrade/total_importsusd.rds"))
mqty <- as.data.table(readRDS("/home/pawan/trade/totaltrade/total_importsqty.rds"))
twodigit <- as.data.table(read_excel("/home/pawan/trade/totaltrade/two-digit.xlsx"))

clean_trade <- function(dt,val_col) {
  dt[,`:=`(HSCode2=substr(HSCode,1,2),HSCode4=substr(HSCode,1,4))]
  dt[,(val_col):=as.numeric(gsub(",","",get(val_col)))]
  dt[is.na(get(val_col)),(val_col):=0]
  dt[,HSCode2:=as.numeric(HSCode2)]; dt }

usd=clean_trade(usd,"Export"); qty=clean_trade(qty,"Export")
musd=clean_trade(musd,"Import"); mqty=clean_trade(mqty,"Import")
twodigit[,HSCode2:=as.numeric(HSCode2)]
qty  <- merge(qty, twodigit, by="HSCode2",all.x=TRUE)
mqty <- merge(mqty,twodigit, by="HSCode2",all.x=TRUE)

exp_usd <- usd[,.(Exp_USD=sum(Export,na.rm=TRUE)),by=Year]
imp_usd <- musd[,.(Imp_USD=sum(Import,na.rm=TRUE)),by=Year]
exp_qty <- qty[,.(Exp_Qty=sum(Export,na.rm=TRUE)),by=Year]
imp_qty <- mqty[,.(Imp_Qty=sum(Import,na.rm=TRUE)),by=Year]
trade <- Reduce(function(a,b) merge(a,b,by="Year"),list(exp_usd,imp_usd,exp_qty,imp_qty))
trade[,`:=`(Balance=Exp_USD-Imp_USD, Coverage=round(Exp_USD*100/Imp_USD,1),
            Exp_Gr=round((Exp_USD/shift(Exp_USD)-1)*100,1),
            Imp_Gr=round((Imp_USD/shift(Imp_USD)-1)*100,1))]
```

## Introduction: India's Trade in Numbers

India crossed the **USD 1 trillion merchandise trade** milestone in 2021–22. Total exports reached approximately USD 422 billion in 2022–23, while imports exceeded USD 714 billion — a trade deficit of **USD 265 billion**, larger than the entire GDP of many middle-income countries. This persistent deficit is structural: it is driven by India's energy import dependence, its under-representation in manufactured goods exports, and its growing import demand for industrial inputs.

---

## The Big Picture: Exports, Imports, and the Deficit

```{r fig-trade-balance}
#| fig-cap: "India's Merchandise Trade: Exports, Imports, and Trade Deficit (USD Billion)"
long_tr <- melt(trade[,.(Year,Exp_USD,Imp_USD)],id="Year",
                variable.name="Flow",value.name="USD")
long_tr[,tip:=paste0(ifelse(Flow=="Exp_USD","<b>Exports</b>","<b>Imports</b>"),
  "<br>Year: ",Year,"<br>USD <b>",format(round(USD/1000),big.mark=","),"B</b>")]
trade[,bal_tip:=paste0("Year: ",Year,"<br>Trade Balance: USD <b>",
  format(round(Balance/1000),big.mark=","),"B</b><br>Coverage ratio: <b>",Coverage,"%</b>")]

p <- ggplot()+
  geom_col_interactive(data=trade,aes(x=Year,y=Balance/1000,tooltip=bal_tip,data_id=Year,
                                       fill=Balance<0),width=0.7,alpha=0.3)+
  geom_line_interactive(data=long_tr,aes(x=Year,y=USD/1000,color=Flow,tooltip=tip,
                                          data_id=paste(Flow,Year),group=Flow),linewidth=1.4)+
  geom_point_interactive(data=long_tr,aes(x=Year,y=USD/1000,color=Flow,tooltip=tip,
                                           data_id=paste(Flow,Year)),size=2.5)+
  geom_hline(yintercept=0,linetype="dashed",color="#7f8c8d")+
  scale_fill_manual(values=c("TRUE"="#e74c3c","FALSE"="#27ae60"),guide="none")+
  scale_color_manual(values=c("Exp_USD"="#27ae60","Imp_USD"="#e74c3c"),
                     labels=c("Exports","Imports"))+
  scale_y_continuous(labels=label_dollar(suffix="B",scale=1))+
  labs(y="USD Billions",x="Year",subtitle="Bars = trade balance; lines = exports & imports",
       caption="Source: Ministry of Commerce | Analysis: Pawan")+my_theme
girafe(ggobj=p,width_svg=12,height_svg=6,options=std_opts)
```

```{r tbl-trade}
#| tbl-cap: "India's Merchandise Trade Summary (USD Billion)"
t_tbl <- trade[,.(Year,Exp_USD=round(Exp_USD/1000,1),Imp_USD=round(Imp_USD/1000,1),
                  Balance=round(Balance/1000,1),Coverage,Exp_Gr,Imp_Gr)]
gt(t_tbl)|>
  tab_header(title=md("**India's Merchandise Trade (USD Billion)**"),
             subtitle="Balance = Exports minus Imports; Coverage = Exports ÷ Imports × 100")|>
  cols_label(Year="Year",Exp_USD="Exports",Imp_USD="Imports",Balance="Balance",
             Coverage="Coverage (%)",Exp_Gr="Export YoY %",Imp_Gr="Import YoY %")|>
  fmt_number(columns=c(Exp_USD,Imp_USD,Balance),decimals=1)|>
  fmt_number(columns=c(Coverage,Exp_Gr,Imp_Gr),decimals=1)|>
  data_color(columns=Balance,palette=c("#fee2e2","#fca5a5","#dc2626","#fff","#86efac","#16a34a"))|>
  tab_style(style=cell_text(color="#e74c3c",weight="bold"),
            locations=cells_body(columns=Exp_Gr,rows=!is.na(Exp_Gr)&Exp_Gr>30))|>
  tab_source_note("Source: Ministry of Commerce, India | Analysis: Pawan")
```

India's **trade coverage ratio** — exports as a percentage of imports — has declined from around 70% in the early 2000s to approximately **55–60%** in recent years, indicating that import growth has consistently outpaced export growth.

---

## Export Growth: Year-on-Year Volatility

```{r fig-growth}
#| fig-cap: "India's Export and Import Growth Rates (YoY %): High Volatility Around Crisis Years"
gr_long <- melt(trade[!is.na(Exp_Gr),.(Year,Exp_Gr,Imp_Gr)],
                id="Year",variable.name="Flow",value.name="Growth")
gr_long[,tip:=paste0(ifelse(Flow=="Exp_Gr","<b>Export</b>","<b>Import</b>"),
  " growth<br>Year: ",Year,"<br>Change: <b>",Growth,"%</b>")]
pg <- ggplot(gr_long,aes(x=Year,y=Growth,fill=Flow,tooltip=tip,data_id=paste(Flow,Year)))+
  geom_col_interactive(position="dodge",width=0.65,alpha=0.88)+
  geom_hline(yintercept=0,color="#2c3e50",linewidth=0.5)+
  scale_fill_manual(values=c("Exp_Gr"="#27ae60","Imp_Gr"="#e74c3c"),
                    labels=c("Export Growth","Import Growth"))+
  scale_y_continuous(labels=label_percent(scale=1))+
  labs(y="Growth (%)",x="Year",subtitle="2008–09 and 2020–21 COVID shocks; 2021–22 commodity boom",
       caption="Source: Ministry of Commerce | Analysis: Pawan")+my_theme
girafe(ggobj=pg,width_svg=11,height_svg=6,options=std_opts)
```

---

## Cereal Exports: India as the World's Rice Granary

```{r fig-cereals}
#| fig-cap: "India's Cereal Exports by Commodity (Thousand Tonnes) — Rice Dominates"
cereals <- qty[HSCode2==10]
cereals[,Commodity:=fcase(HSCode4=="1006","Rice",HSCode4=="1001","Wheat",
  HSCode4=="1005","Maize",HSCode4=="1008","Millet & Buckwheat",default="Other Cereals")]
cer_yr <- cereals[,.(Qty=sum(Export,na.rm=TRUE)),by=.(Year,Commodity)]
cer_yr[,Total:=sum(Qty),by=Year][,Share:=round(Qty*100/Total)]
cer_yr[,Commodity:=factor(Commodity,
  levels=c("Other Cereals","Millet & Buckwheat","Maize","Wheat","Rice"))]
cer_yr[,tip:=paste0("<b>",Commodity,"</b><br>Year: ",Year,"<br>",
  "Qty: ",format(round(Qty),big.mark=",")," '000T<br>Share: ",Share,"%")]

pc <- ggplot(cer_yr,aes(x=Year,y=Qty,fill=Commodity,tooltip=tip,data_id=paste(Commodity,Year)))+
  geom_bar_interactive(stat="identity",position="stack",width=0.75,alpha=0.9)+
  geom_text(aes(label=ifelse(Share>=8,paste0(Share,"%"),"")),
            position=position_stack(vjust=0.5),size=2.8,color="white",fontface="bold")+
  scale_fill_viridis(discrete=TRUE,option="D")+
  scale_y_continuous(labels=comma_format())+
  labs(y="Quantity ('000 Tonnes)",x="Year",
       subtitle="Rice consistently >60% of cereal exports; wheat is policy-driven",
       caption="Source: Ministry of Commerce | Analysis: Pawan")+my_theme
girafe(ggobj=pc,width_svg=11,height_svg=6,options=std_opts)
```

```{r tbl-cereals}
#| tbl-cap: "India's Cereal Export Volumes by Commodity (Thousand Tonnes)"
cw <- dcast(cer_yr,Year~Commodity,value.var="Qty",fun.aggregate=sum)
cw[,Total:=rowSums(.SD,na.rm=TRUE),.SDcols=setdiff(names(cw),"Year")]
cw[,Rice_Share:=paste0(round(Rice*100/Total),"%")]
gt(cw[,.(Year,Rice,Wheat,Maize,`Millet & Buckwheat`,`Other Cereals`,Total,Rice_Share)])|>
  tab_header(title=md("**Cereal Export Volumes ('000 Tonnes)**"))|>
  fmt_number(columns=c(Rice,Wheat,Maize,`Millet & Buckwheat`,`Other Cereals`,Total),
             decimals=0,use_seps=TRUE)|>
  cols_label(Rice_Share="Rice %")|>
  data_color(columns=Rice,palette=c("#f0fdf4","#86efac","#16a34a"))|>
  tab_source_note("Source: Ministry of Commerce | Analysis: Pawan")
```

India is the **world's largest rice exporter**, accounting for ~40% of global rice trade. The data shows wheat exports peaking in 2021–22 when global prices were high, before being sharply curtailed by an export ban in May 2022 to protect domestic supplies.

---

## Petroleum Imports: India's Energy Anchor

```{r fig-petro}
#| fig-cap: "India's Petroleum & Mineral Fuel Imports by Type (Thousand Tonnes)"
petro <- mqty[HSCode2==27]
petro[,Commodity:=fcase(HSCode4=="2709","Crude Petroleum",HSCode4=="2701","Coal",
                         HSCode4=="2711","Petroleum Gases",default="Other Mineral Fuels")]
pet_yr <- petro[,.(Qty=sum(Import,na.rm=TRUE)),by=.(Year,Commodity)]
pet_yr[,Total:=sum(Qty),by=Year][,Share:=round(Qty*100/Total)]
pet_yr[,tip:=paste0("<b>",Commodity,"</b><br>Year: ",Year,"<br>",
  "Qty: ",format(round(Qty),big.mark=",")," '000T<br>Share: ",Share,"%")]

pp <- ggplot(pet_yr,aes(x=Year,y=Qty,fill=Commodity,tooltip=tip,data_id=paste(Commodity,Year)))+
  geom_bar_interactive(stat="identity",position="stack",width=0.75,alpha=0.9)+
  geom_text(aes(label=ifelse(Share>=8,paste0(Share,"%"),"")),
            position=position_stack(vjust=0.5),size=2.8,color="white",fontface="bold")+
  scale_fill_manual(values=c("Crude Petroleum"="#e67e22","Coal"="#2c3e50",
                             "Petroleum Gases"="#8e44ad","Other Mineral Fuels"="#bdc3c7"))+
  scale_y_continuous(labels=comma_format())+
  labs(y="Quantity ('000 Tonnes)",x="Year",
       subtitle="Crude petroleum dominates; LNG imports rising with gas infrastructure",
       caption="Source: Ministry of Commerce | Analysis: Pawan")+my_theme
girafe(ggobj=pp,width_svg=11,height_svg=6,options=std_opts)
```

Petroleum (HS Chapter 27) is India's single largest import category, constituting **25–30% of total import value**. India's crude oil import dependence (85% of consumption from overseas) makes every USD 10/bbl increase in crude prices add approximately **USD 14–15 billion** to the annual import bill.

---

## Fertilizer Imports: The Agricultural Dependence

```{r fig-fert-trade}
#| fig-cap: "India's Fertilizer Import Volumes by Type (Thousand Tonnes)"
fert <- mqty[HSCode2==31]
fert[,Commodity:=fcase(HSCode4=="3102","Nitrogenous (Urea)",HSCode4=="3104","Potassic (MOP)",
                        HSCode4=="3105","DAP & NPK",default="Other Fertilizers")]
fert_yr <- fert[,.(Qty=sum(Import,na.rm=TRUE)),by=.(Year,Commodity)]
fert_yr[,Total:=sum(Qty),by=Year][,Share:=round(Qty*100/Total)]
fert_yr[,tip:=paste0("<b>",Commodity,"</b><br>Year: ",Year,"<br>",
  "Qty: ",format(round(Qty),big.mark=",")," '000T<br>Share: ",Share,"%")]

pf <- ggplot(fert_yr,aes(x=Year,y=Qty,fill=Commodity,tooltip=tip,data_id=paste(Commodity,Year)))+
  geom_bar_interactive(stat="identity",position="stack",width=0.75,alpha=0.9)+
  geom_text(aes(label=ifelse(Share>=8,paste0(Share,"%"),"")),
            position=position_stack(vjust=0.5),size=2.8,color="white",fontface="bold")+
  scale_fill_viridis(discrete=TRUE,option="C")+
  scale_y_continuous(labels=comma_format())+
  labs(y="Quantity ('000 Tonnes)",x="Year",
       caption="Source: Ministry of Commerce | Analysis: Pawan")+my_theme
girafe(ggobj=pf,width_svg=11,height_svg=6,options=std_opts)
```

---

## Edible Oil Imports: Volume Trajectory

```{r fig-oils-trade}
#| fig-cap: "India's Edible Oil & Fat Import Volume (HS Chapter 15)"
oils <- mqty[HSCode2==15]
oils_yr <- oils[,.(Qty=sum(Import,na.rm=TRUE)),by=Year]
oils_yr[,tip:=paste0("Year: <b>",Year,"</b><br>Edible oils import: ",
  format(round(Qty),big.mark=",")," '000T")]
po <- ggplot(oils_yr,aes(x=Year,y=Qty,tooltip=tip,data_id=Year,group=1))+
  geom_area_interactive(fill="#f39c12",alpha=0.25)+
  geom_line_interactive(linewidth=1.3,color="#d68910")+
  geom_point_interactive(size=2.5,color="#b7770d")+
  scale_y_continuous(labels=comma_format(suffix=" '000T"))+
  labs(y="Quantity ('000 Tonnes)",x="Year",
       caption="Source: Ministry of Commerce | Analysis: Pawan")+my_theme
girafe(ggobj=po,width_svg=10,height_svg=5,options=std_opts)
```

---

## Conclusion

The trade data paints a clear picture: India has become a significant global trader, but its import basket remains dominated by three categories where structural dependence is high — **petroleum, fertilizers, and edible oils**. These three together account for over 40% of total import value and are all subject to large geopolitical and commodity price risks. Meanwhile, export growth has been led by petroleum products (from refinery capacity), engineering goods, and chemicals — but manufactured goods' share remains well below India's aspiration to become a leading manufacturing exporter. Closing this gap is the central trade policy challenge of the coming decade.

---
*Data: Ministry of Commerce, India; DGCI&S. Analysis: Pawan.*
