---
title: "Basic Statistics"
subtitle: "Concepts, Formulas, and Applications in R"
author: "Pawan"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    highlight-style: github
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

**Basic statistics** is the foundation of all data analysis. It provides tools to summarize, describe, and understand data before performing any advanced modelling. It answers the most fundamental question: *What does the data look like?*

---

# Key Concepts and Definitions

## 1. Population vs Sample

- **Population**: The complete set of all items of interest (e.g., all students in India).
- **Sample**: A subset of the population selected for study (e.g., 500 students from Delhi).

We use **sample statistics** to estimate **population parameters**.

| Measure | Population Symbol | Sample Symbol |
|---------|------------------|---------------|
| Mean | μ (mu) | x̄ |
| Standard Deviation | σ (sigma) | s |
| Size | N | n |

---

## 2. Measures of Central Tendency

These describe the **centre** of the data.

### Mean (Arithmetic Average)

$$\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n}$$

### Median

The **middle value** when data is sorted. For $n$ observations:

- If $n$ is odd: Median = $\left(\frac{n+1}{2}\right)$th value
- If $n$ is even: Median = average of $\left(\frac{n}{2}\right)$th and $\left(\frac{n}{2}+1\right)$th values

### Mode

The value that **occurs most frequently** in the dataset.

---

## 3. Measures of Dispersion

These describe how **spread out** the data is.

### Variance

$$s^2 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n-1}$$

### Standard Deviation

$$s = \sqrt{\frac{\sum_{i=1}^{n}(x_i - \bar{x})^2}{n-1}}$$

### Range

$$\text{Range} = x_{\max} - x_{\min}$$

### Interquartile Range (IQR)

$$\text{IQR} = Q_3 - Q_1$$

where $Q_1$ = 25th percentile and $Q_3$ = 75th percentile.

---

## 4. Measures of Shape

### Skewness

Measures **asymmetry** of the distribution.

$$\text{Skewness} = \frac{1}{n} \sum_{i=1}^{n} \left(\frac{x_i - \bar{x}}{s}\right)^3$$

- Skewness = 0 → Symmetric
- Skewness > 0 → Right-skewed (tail on the right)
- Skewness < 0 → Left-skewed (tail on the left)

### Kurtosis

Measures **peakedness** of the distribution.

$$\text{Kurtosis} = \frac{1}{n} \sum_{i=1}^{n} \left(\frac{x_i - \bar{x}}{s}\right)^4$$

---

# When and Where to Use Basic Statistics

| Situation | Tool |
|-----------|------|
| Summarising exam scores | Mean, Median, SD |
| Comparing income distributions | Median, IQR (robust to outliers) |
| Understanding data shape | Skewness, Kurtosis |
| Checking data spread | Variance, Range |
| Exploratory Data Analysis (EDA) | All of the above |

**Real-world examples:**
- A hospital computes the **mean** age of patients to plan resources.
- An economist uses the **median** income (not mean) because income data is right-skewed.
- A teacher computes **standard deviation** of marks to judge class variability.

---

# Implementation in R

## Loading and Exploring Data

```{r}
# Use the built-in 'mtcars' dataset
data(mtcars)

# First look at the data
head(mtcars)
str(mtcars)
```

## Measures of Central Tendency

```{r}
# Focus on miles per gallon (mpg)
x <- mtcars$mpg

cat("Mean    :", mean(x), "\n")
cat("Median  :", median(x), "\n")
cat("Mode    :", as.numeric(names(sort(table(x), decreasing = TRUE)[1])), "\n")
```

## Measures of Dispersion

```{r}
cat("Variance:", var(x), "\n")
cat("Std Dev :", sd(x), "\n")
cat("Range   :", range(x), "\n")
cat("IQR     :", IQR(x), "\n")
```

## Quantiles and Summary

```{r}
# Five-number summary + mean
summary(x)

# Custom quantiles
quantile(x, probs = c(0.25, 0.50, 0.75, 0.90))
```

## Skewness and Kurtosis

```{r}
# Install e1071 if not available: install.packages("e1071")
library(e1071)

cat("Skewness:", skewness(x), "\n")
cat("Kurtosis:", kurtosis(x), "\n")
```

## Full Descriptive Statistics Table

```{r}
# A clean descriptive statistics summary
desc_stats <- data.frame(
  Statistic = c("N", "Mean", "Median", "SD", "Variance",
                "Min", "Max", "Range", "IQR", "Skewness", "Kurtosis"),
  Value = round(c(
    length(x), mean(x), median(x), sd(x), var(x),
    min(x), max(x), diff(range(x)), IQR(x),
    skewness(x), kurtosis(x)
  ), 3)
)

print(desc_stats)
```

## Visualisation

```{r}
#| fig-width: 10
#| fig-height: 4

par(mfrow = c(1, 3))

# Histogram
hist(x,
     main = "Distribution of MPG",
     xlab = "Miles Per Gallon",
     col = "steelblue", border = "white",
     breaks = 10)
abline(v = mean(x), col = "red", lwd = 2, lty = 2)
abline(v = median(x), col = "darkgreen", lwd = 2, lty = 2)
legend("topright", legend = c("Mean", "Median"),
       col = c("red", "darkgreen"), lty = 2, lwd = 2)

# Boxplot
boxplot(x,
        main = "Boxplot of MPG",
        ylab = "Miles Per Gallon",
        col = "lightblue", border = "navy")

# QQ Plot (normality check)
qqnorm(x, main = "Normal Q-Q Plot")
qqline(x, col = "red", lwd = 2)
```

---

# Interpretation

Looking at the MPG data from `mtcars`:

- The **mean MPG** (~20.09) is slightly higher than the **median** (~19.2), suggesting a mild right skew.
- The **standard deviation** (~6.03) tells us most cars fall within ±6 MPG of the mean.
- The **IQR** spans from ~15.4 to ~22.8 MPG, covering the middle 50% of cars.
- The **Q-Q plot** helps assess if the data follows a normal distribution.

---

# Summary

> Basic statistics transforms raw numbers into meaningful insights. Always start any data analysis by computing and visualising these descriptive measures — they guide every decision that follows.

| What | Formula | R Function |
|------|---------|-----------|
| Mean | $\bar{x} = \frac{\sum x_i}{n}$ | `mean()` |
| Median | Middle value | `median()` |
| Variance | $s^2 = \frac{\sum(x_i - \bar{x})^2}{n-1}$ | `var()` |
| Std Dev | $s = \sqrt{s^2}$ | `sd()` |
| IQR | $Q_3 - Q_1$ | `IQR()` |
| Summary | All at once | `summary()` |
