---
title: "The Central Limit Theorem"
subtitle: "The Most Important Theorem in Statistics"
author: "Pawan"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    highlight-style: github
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

The **Central Limit Theorem (CLT)** is arguably the most important result in all of statistics. It explains *why* the normal distribution appears everywhere in practice and forms the theoretical backbone of hypothesis testing, confidence intervals, and much of inferential statistics.

> "No matter what shape the original population has, the distribution of the sample mean becomes approximately normal as the sample size increases."

---

# The Theorem — Formal Statement

Let $X_1, X_2, \ldots, X_n$ be a random sample of size $n$ drawn from **any** population with:

- Mean: $\mu$
- Finite variance: $\sigma^2$

Then as $n \to \infty$, the distribution of the **sample mean** $\bar{X}$ approaches a normal distribution:

$$\bar{X} \sim N\left(\mu,\ \frac{\sigma^2}{n}\right)$$

Or equivalently, the **standardised** sample mean:

$$Z = \frac{\bar{X} - \mu}{\sigma / \sqrt{n}} \xrightarrow{d} N(0, 1)$$

converges in distribution to the standard normal.

---

# Key Components

## Standard Error (SE)

The standard deviation of the sampling distribution of the mean is called the **Standard Error**:

$$SE = \frac{\sigma}{\sqrt{n}}$$

- As $n$ increases → SE **decreases** → sample mean is more precise
- Doubling sample size reduces SE by factor of $\sqrt{2} \approx 1.41$

## Rule of Thumb for Sample Size

| Population Shape | Minimum $n$ for CLT |
|-----------------|---------------------|
| Roughly symmetric | $n \geq 30$ |
| Moderately skewed | $n \geq 50$ |
| Heavily skewed / extreme outliers | $n \geq 100$ |

---

# Why Is the CLT So Important?

1. **Inference without knowing population distribution**: We can build confidence intervals and run t-tests even when the population is not normal — as long as $n$ is large enough.

2. **Foundation of hypothesis testing**: The Z-test and t-test are direct applications of the CLT.

3. **Quality control**: Manufacturing uses CLT to monitor process averages via control charts (X-bar charts).

4. **Finance**: Portfolio returns (averages of many individual stock returns) are approximately normal.

5. **Polling and surveys**: Sample proportions (a type of mean) follow a normal distribution for large samples.

---

# When and Where to Use the CLT

| Application | Description |
|-------------|-------------|
| Confidence intervals | $\bar{x} \pm z_{\alpha/2} \cdot \frac{s}{\sqrt{n}}$ |
| One-sample Z-test | Test if population mean equals a value |
| Polling | Estimate proportions from survey data |
| Quality control (SPC) | Monitor sample means over time |
| Simulation / bootstrapping | Theoretical justification |

---

# Implementation in R

## Simulating the CLT — Uniform Distribution

We draw repeated samples from a **Uniform distribution** (clearly non-normal) and watch the distribution of sample means converge to normal.

```{r}
set.seed(42)

# Parameters
pop_mean <- 0.5   # mean of Uniform(0,1)
pop_sd   <- sqrt(1/12)  # sd of Uniform(0,1)

# Sample sizes to demonstrate
sample_sizes <- c(1, 5, 30, 100)
n_simulations <- 5000

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

for (n in sample_sizes) {
  # Draw 5000 sample means, each from a sample of size n
  sample_means <- replicate(n_simulations, mean(runif(n, 0, 1)))

  hist(sample_means,
       breaks = 40,
       probability = TRUE,
       col = "steelblue",
       border = "white",
       main = paste("n =", n),
       xlab = "Sample Mean",
       xlim = c(0, 1))

  # Overlay the theoretical normal curve
  curve(dnorm(x, mean = pop_mean, sd = pop_sd / sqrt(n)),
        add = TRUE, col = "red", lwd = 2.5)
}
```

## Simulating the CLT — Exponential Distribution (Very Skewed)

```{r}
set.seed(123)

# Exponential with rate = 1: mean = 1, sd = 1
sample_sizes2 <- c(1, 10, 30, 100)

par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

for (n in sample_sizes2) {
  sample_means <- replicate(n_simulations, mean(rexp(n, rate = 1)))

  hist(sample_means,
       breaks = 50,
       probability = TRUE,
       col = "coral",
       border = "white",
       main = paste("Exponential — n =", n),
       xlab = "Sample Mean")

  # Theoretical normal overlay
  curve(dnorm(x, mean = 1, sd = 1 / sqrt(n)),
        add = TRUE, col = "navy", lwd = 2.5)
}
```

## Demonstrating the Standard Error

```{r}
# True SE shrinks as n increases
n_values <- c(10, 30, 50, 100, 200, 500, 1000)
true_se   <- 1 / sqrt(n_values)  # for Exp(1), sigma = 1

data.frame(
  n  = n_values,
  SE = round(true_se, 4)
)
```

```{r}
plot(n_values, true_se,
     type = "b", pch = 16, col = "steelblue", lwd = 2,
     xlab = "Sample Size (n)",
     ylab = "Standard Error",
     main = "Standard Error Decreases as n Increases")
abline(h = 0, lty = 2, col = "grey")
```

## Practical Application — Confidence Interval

Suppose we want to estimate the mean weight of students at a university. We take a sample of $n = 50$ students.

```{r}
set.seed(99)

# Simulated weight data (population: skewed, like real weight data)
weights <- rexp(50, rate = 1/70) + 40   # mean ≈ 110 lbs, shifted

n    <- length(weights)
xbar <- mean(weights)
s    <- sd(weights)
se   <- s / sqrt(n)

# 95% Confidence Interval using CLT (z = 1.96)
z    <- qnorm(0.975)
ci_lower <- xbar - z * se
ci_upper <- xbar + z * se

cat(sprintf("Sample Mean   : %.2f\n", xbar))
cat(sprintf("Std Error     : %.2f\n", se))
cat(sprintf("95%% CI        : (%.2f, %.2f)\n", ci_lower, ci_upper))
```

The 95% CI means: if we repeated this sampling procedure 100 times, approximately 95 of those intervals would contain the true population mean.

## Verifying Coverage Probability

```{r}
set.seed(42)

# True mean = 70 (Exponential with rate=1/70, shifted)
true_mean <- 110
coverage_count <- 0
n_trials <- 1000
n_each <- 50

for (i in 1:n_trials) {
  samp <- rexp(n_each, rate = 1/70) + 40
  xbar_i <- mean(samp)
  se_i   <- sd(samp) / sqrt(n_each)
  lower  <- xbar_i - 1.96 * se_i
  upper  <- xbar_i + 1.96 * se_i
  if (lower <= true_mean && true_mean <= upper) {
    coverage_count <- coverage_count + 1
  }
}

cat(sprintf("Empirical coverage: %.1f%% (target: 95%%)\n",
            100 * coverage_count / n_trials))
```

---

# Summary

The Central Limit Theorem tells us:

1. **Any** population + **large enough** sample → sample mean is **approximately normal**
2. The sampling distribution centres at $\mu$ with spread $\sigma/\sqrt{n}$
3. Larger $n$ → smaller standard error → more precise estimates

| Component | Formula | R function |
|-----------|---------|-----------|
| Sample Mean | $\bar{x} = \frac{\sum x_i}{n}$ | `mean()` |
| Standard Error | $SE = s/\sqrt{n}$ | `sd(x)/sqrt(n)` |
| 95% CI | $\bar{x} \pm 1.96 \cdot SE$ | `t.test(x)` |
| Z-score | $Z = (\bar{x} - \mu)/SE$ | manual computation |
| Normal quantile | $z_{\alpha/2}$ | `qnorm(0.975)` |

> The CLT is why we can use normal-based methods on real-world data — even when the raw data is not normal.
