---
title: "Regression Analysis"
subtitle: "Simple and Multiple Linear Regression in R"
author: "Pawan"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    highlight-style: github
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

**Regression analysis** is a statistical method used to understand and quantify the **relationship between variables**. It helps us:

- Explain how a dependent variable changes with one or more independent variables.
- **Predict** future values of the dependent variable.

> Example: How does a car's fuel efficiency (MPG) change with its weight and horsepower?

---

# Types of Regression

| Type | Response Variable | Predictors |
|------|-----------------|------------|
| Simple Linear | Continuous | 1 continuous |
| Multiple Linear | Continuous | 2+ continuous/categorical |
| Logistic | Binary (0/1) | Any |
| Polynomial | Continuous | Powered terms |

This paper focuses on **Simple and Multiple Linear Regression**.

---

# Simple Linear Regression

## Definition

Models the linear relationship between one predictor $X$ and a response $Y$:

$$Y = \beta_0 + \beta_1 X + \varepsilon$$

where:

- $Y$ = dependent variable (response)
- $X$ = independent variable (predictor)
- $\beta_0$ = intercept (value of $Y$ when $X = 0$)
- $\beta_1$ = slope (change in $Y$ per unit change in $X$)
- $\varepsilon \sim N(0, \sigma^2)$ = random error term

## Ordinary Least Squares (OLS) Estimation

OLS minimises the **sum of squared residuals**:

$$\text{SSR} = \sum_{i=1}^{n}(y_i - \hat{y}_i)^2 = \sum_{i=1}^{n}(y_i - \hat{\beta}_0 - \hat{\beta}_1 x_i)^2$$

The closed-form solutions are:

$$\hat{\beta}_1 = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^{n}(x_i - \bar{x})^2} = \frac{S_{xy}}{S_{xx}}$$

$$\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}$$

## Goodness of Fit — R-squared

$$R^2 = 1 - \frac{\text{SSRes}}{\text{SST}} = 1 - \frac{\sum(y_i - \hat{y}_i)^2}{\sum(y_i - \bar{y})^2}$$

- $R^2 = 1$: Perfect fit
- $R^2 = 0$: Model explains nothing
- $0 < R^2 < 1$: Proportion of variance explained

---

# Multiple Linear Regression

## Definition

$$Y = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k + \varepsilon$$

In matrix notation:

$$\mathbf{Y} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon}$$

OLS solution:

$$\hat{\boldsymbol{\beta}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y}$$

## Adjusted R-squared

For multiple regression, use **Adjusted $R^2$** to penalise for extra predictors:

$$\bar{R}^2 = 1 - \frac{(1-R^2)(n-1)}{n-k-1}$$

where $k$ = number of predictors and $n$ = sample size.

---

# Assumptions of Linear Regression (LINE)

| Letter | Assumption |
|--------|-----------|
| **L** | **L**inearity: relationship between $X$ and $Y$ is linear |
| **I** | **I**ndependence: observations are independent |
| **N** | **N**ormality: residuals are normally distributed |
| **E** | **E**qual variance: residuals have constant variance (homoscedasticity) |

---

# When and Where to Use

| Scenario | Use Case |
|----------|----------|
| Economics | Predicting GDP from investment and consumption |
| Medicine | Predicting blood pressure from age and BMI |
| Real estate | Predicting house price from area and location |
| Agriculture | Predicting crop yield from rainfall and temperature |
| HR analytics | Predicting salary from experience and education |

---

# Implementation in R

## Simple Linear Regression

```{r}
# Use mtcars: predict mpg from weight (wt)
data(mtcars)

# Fit the model
model_simple <- lm(mpg ~ wt, data = mtcars)
summary(model_simple)
```

### Interpretation of Output

```{r}
cat("Intercept (β₀)  :", coef(model_simple)[1], "\n")
cat("Slope (β₁)      :", coef(model_simple)[2], "\n")
cat("R-squared       :", summary(model_simple)$r.squared, "\n")
cat("Adj R-squared   :", summary(model_simple)$adj.r.squared, "\n")
```

The slope of ~−5.34 means: for every 1000 lb increase in weight, MPG decreases by ~5.34.

### Scatter Plot with Regression Line

```{r}
#| fig-width: 7
#| fig-height: 5

plot(mtcars$wt, mtcars$mpg,
     pch = 16, col = "steelblue",
     xlab = "Weight (1000 lbs)",
     ylab = "Miles Per Gallon",
     main = "Simple Linear Regression: MPG ~ Weight")

abline(model_simple, col = "red", lwd = 2)

# Add R² label
r2 <- round(summary(model_simple)$r.squared, 3)
legend("topright", legend = paste("R² =", r2), bty = "n")
```

### Residual Diagnostics

```{r}
#| fig-width: 10
#| fig-height: 8

par(mfrow = c(2, 2))
plot(model_simple, col = "steelblue", pch = 16)
```

The four diagnostic plots check:

1. **Residuals vs Fitted**: Should show no pattern (linearity, homoscedasticity)
2. **Normal Q-Q**: Points should follow the diagonal line (normality)
3. **Scale-Location**: Should be horizontal (equal variance)
4. **Residuals vs Leverage**: Identifies influential points

## Multiple Linear Regression

```{r}
# Predict mpg from weight AND horsepower
model_multi <- lm(mpg ~ wt + hp, data = mtcars)
summary(model_multi)
```

```{r}
cat("Intercept        :", round(coef(model_multi)[1], 3), "\n")
cat("β₁ (wt)          :", round(coef(model_multi)[2], 3), "\n")
cat("β₂ (hp)          :", round(coef(model_multi)[3], 3), "\n")
cat("Adjusted R²      :", round(summary(model_multi)$adj.r.squared, 3), "\n")
```

Compared to the simple model, adding horsepower increases Adjusted R² — meaning it adds explanatory power.

## Comparing Models

```{r}
# ANOVA test to compare models
anova(model_simple, model_multi)
```

A significant p-value means the more complex model fits significantly better.

## Prediction

```{r}
# Predict MPG for a new car: wt = 3.0, hp = 110
new_car <- data.frame(wt = 3.0, hp = 110)

# Point prediction
pred <- predict(model_multi, newdata = new_car)
cat("Predicted MPG:", round(pred, 2), "\n")

# With 95% prediction interval
pred_interval <- predict(model_multi, newdata = new_car, interval = "prediction")
print(pred_interval)
```

## Checking Multicollinearity (VIF)

```{r}
# install.packages("car")
library(car)
vif(model_multi)
```

- VIF < 5: No serious multicollinearity
- VIF > 10: High multicollinearity — consider removing or combining predictors

## Full Model with All Predictors

```{r}
model_full <- lm(mpg ~ ., data = mtcars)
summary(model_full)$adj.r.squared
```

```{r}
# Stepwise selection to find the best model
model_step <- step(model_full, direction = "backward", trace = 0)
summary(model_step)
```

---

# Coefficient of Determination Summary

```{r}
models <- list(
  "Simple (wt only)"    = model_simple,
  "Multiple (wt + hp)"  = model_multi,
  "Stepwise selected"   = model_step
)

results <- sapply(models, function(m) {
  c(R2     = round(summary(m)$r.squared, 4),
    AdjR2  = round(summary(m)$adj.r.squared, 4),
    AIC    = round(AIC(m), 2))
})

as.data.frame(t(results))
```

---

# Summary

| Concept | Formula | R Function |
|---------|---------|-----------|
| Fit model | $Y = \beta_0 + \beta_1 X$ | `lm(Y ~ X)` |
| View results | — | `summary(model)` |
| Coefficients | $\hat{\boldsymbol{\beta}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y}$ | `coef(model)` |
| Predict | $\hat{y} = \hat{\beta}_0 + \hat{\beta}_1 x$ | `predict(model, newdata)` |
| Diagnostics | — | `plot(model)` |
| VIF | — | `car::vif(model)` |
| Model comparison | — | `anova(m1, m2)`, `AIC()` |

> Regression is the workhorse of statistical modelling. Always check assumptions via diagnostic plots before interpreting coefficients.
