---
title: "Time Series Models"
subtitle: "Analysis, Decomposition, and Forecasting in R"
author: "Pawan"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    highlight-style: github
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

A **time series** is a sequence of data points collected at successive points in time, typically at equally spaced intervals. Time series analysis is used to understand past behaviour and **forecast future values**.

> Examples: monthly rainfall, daily stock prices, quarterly GDP, weekly sales figures, annual temperature records.

---

# Components of a Time Series

Every time series can be decomposed into four components:

| Component | Symbol | Description |
|-----------|--------|-------------|
| **Trend** ($T_t$) | Long-run direction | Rising, falling, or flat over time |
| **Seasonality** ($S_t$) | Regular periodic pattern | Repeats within a known period (month, quarter, year) |
| **Cyclical** ($C_t$) | Long-term wave | Multi-year business/economic cycles |
| **Irregular/Noise** ($I_t$) | Random fluctuation | Unexplained random variation |

## Decomposition Models

**Additive model** (when seasonal variation is constant):

$$Y_t = T_t + S_t + C_t + I_t$$

**Multiplicative model** (when seasonal variation grows with trend):

$$Y_t = T_t \times S_t \times C_t \times I_t$$

---

# Key Concepts and Definitions

## Stationarity

A time series is **stationary** if its statistical properties (mean, variance, autocovariance) do not change over time.

- **Mean stationarity**: $E[Y_t] = \mu$ (constant)
- **Variance stationarity**: $\text{Var}(Y_t) = \sigma^2$ (constant)
- **Covariance stationarity**: $\text{Cov}(Y_t, Y_{t+k})$ depends only on lag $k$, not time $t$

Most real-world series are **non-stationary** and must be transformed.

## Autocorrelation Function (ACF)

Measures the correlation of a series with its own past values:

$$\rho_k = \frac{\text{Cov}(Y_t, Y_{t-k})}{\text{Var}(Y_t)} = \frac{\gamma_k}{\gamma_0}$$

## Partial Autocorrelation Function (PACF)

Measures the correlation between $Y_t$ and $Y_{t-k}$ **after removing** the effects of lags $1, 2, \ldots, k-1$.

---

# The ARIMA Family of Models

## AR(p) — Autoregressive Model

Current value depends on its own past $p$ values:

$$Y_t = \phi_1 Y_{t-1} + \phi_2 Y_{t-2} + \cdots + \phi_p Y_{t-p} + \varepsilon_t$$

## MA(q) — Moving Average Model

Current value depends on past $q$ error terms:

$$Y_t = \varepsilon_t + \theta_1 \varepsilon_{t-1} + \theta_2 \varepsilon_{t-2} + \cdots + \theta_q \varepsilon_{t-q}$$

## ARMA(p, q)

Combination of AR and MA:

$$Y_t = \phi_1 Y_{t-1} + \cdots + \phi_p Y_{t-p} + \varepsilon_t + \theta_1 \varepsilon_{t-1} + \cdots + \theta_q \varepsilon_{t-q}$$

## ARIMA(p, d, q)

When the series is non-stationary, we apply **differencing** $d$ times:

$$\Delta^d Y_t = \text{ARMA}(p, q)$$

where $\Delta Y_t = Y_t - Y_{t-1}$ (first difference).

### Model Selection Guide

| ACF Pattern | PACF Pattern | Suggested Model |
|-------------|--------------|----------------|
| Cuts off after lag $q$ | Tails off | MA($q$) |
| Tails off | Cuts off after lag $p$ | AR($p$) |
| Both tail off | Both tail off | ARMA($p$, $q$) |
| Both non-stationary | — | Difference first, then check |

---

# When and Where to Use Time Series Models

| Domain | Application |
|--------|-------------|
| Economics | GDP forecasting, inflation modelling |
| Finance | Stock price prediction, volatility modelling |
| Meteorology | Rainfall, temperature prediction |
| Retail | Sales forecasting, inventory planning |
| Healthcare | Disease incidence forecasting |
| Energy | Electricity demand forecasting |

---

# Implementation in R

## Loading a Time Series

```{r}
# Built-in dataset: monthly airline passengers (1949-1960)
data("AirPassengers")

# Inspect structure
print(AirPassengers)
class(AirPassengers)
frequency(AirPassengers)  # 12 = monthly
start(AirPassengers)
end(AirPassengers)
```

## Plotting the Series

```{r}
#| fig-width: 9
#| fig-height: 4

plot(AirPassengers,
     main = "Monthly International Airline Passengers (1949–1960)",
     ylab = "Number of Passengers (thousands)",
     xlab = "Year",
     col = "steelblue", lwd = 1.5)
```

Clearly shows an **upward trend** and **growing seasonal pattern** → multiplicative decomposition is appropriate.

## Decomposition

```{r}
#| fig-width: 9
#| fig-height: 7

# Multiplicative decomposition (seasonal variation grows with level)
decomp <- decompose(AirPassengers, type = "multiplicative")
plot(decomp)
```

```{r}
# View seasonal indices
round(decomp$seasonal[1:12], 4)
```

## Stationarity Testing — Augmented Dickey-Fuller Test

$$H_0: \text{Series has a unit root (non-stationary)}$$
$$H_1: \text{Series is stationary}$$

```{r}
# install.packages("tseries")
library(tseries)

adf.test(AirPassengers)
```

If p-value > 0.05 → fail to reject $H_0$ → series is non-stationary → need differencing.

## ACF and PACF Plots

```{r}
#| fig-width: 10
#| fig-height: 4

# Log-transform to stabilise variance, then first-difference
log_ap <- log(AirPassengers)
diff_log_ap <- diff(log_ap, differences = 1)

par(mfrow = c(1, 2))
acf(diff_log_ap,  main = "ACF of Differenced Log Passengers",  lag.max = 36)
pacf(diff_log_ap, main = "PACF of Differenced Log Passengers", lag.max = 36)
```

Spikes at seasonal lags (12, 24) indicate seasonal ARIMA is needed.

## Fitting ARIMA Model

```{r}
# Manual ARIMA fit
model_arima <- arima(log_ap,
                     order = c(0, 1, 1),          # non-seasonal: d=1, MA(1)
                     seasonal = list(order = c(0, 1, 1), period = 12))

summary(model_arima)
```

## Auto-ARIMA (automatic order selection)

```{r}
# install.packages("forecast")
library(forecast)

model_auto <- auto.arima(AirPassengers, lambda = 0)  # lambda=0 for log transform
print(model_auto)
```

## Forecasting

```{r}
#| fig-width: 9
#| fig-height: 5

# Forecast 24 months ahead
fc <- forecast(model_auto, h = 24)

plot(fc,
     main = "24-Month Forecast: Airline Passengers",
     ylab = "Passengers (thousands)",
     xlab = "Year",
     col = "steelblue")
```

```{r}
# View forecast values with confidence intervals
print(fc)
```

## Checking Residuals

```{r}
#| fig-width: 9
#| fig-height: 5

# Residuals should look like white noise
checkresiduals(model_auto)
```

A good model leaves **white noise residuals**:
- No autocorrelation in residuals
- Residuals approximately normally distributed
- Ljung-Box test p-value > 0.05

## Accuracy Metrics

```{r}
accuracy(model_auto)
```

| Metric | Description |
|--------|-------------|
| **ME** | Mean Error |
| **RMSE** | Root Mean Squared Error |
| **MAE** | Mean Absolute Error |
| **MAPE** | Mean Absolute Percentage Error |
| **MASE** | Mean Absolute Scaled Error |

Lower values = better model fit.

## Exponential Smoothing (ETS) — Alternative Approach

```{r}
model_ets <- ets(AirPassengers)
summary(model_ets)

fc_ets <- forecast(model_ets, h = 24)
accuracy(fc_ets)
```

ETS models are often competitive with ARIMA and are simpler to interpret.

---

# Summary

| Concept | Formula / Description | R Function |
|---------|----------------------|-----------|
| Create ts object | `ts(data, start, frequency)` | `ts()` |
| Plot | — | `plot()` |
| Decompose | $Y_t = T_t + S_t + I_t$ | `decompose()` |
| Stationarity test | ADF test | `tseries::adf.test()` |
| ACF / PACF | $\rho_k = \gamma_k / \gamma_0$ | `acf()`, `pacf()` |
| Fit ARIMA | $\text{ARIMA}(p,d,q)$ | `arima()` |
| Auto-fit | — | `forecast::auto.arima()` |
| Forecast | — | `forecast()` |
| Check residuals | Ljung-Box test | `checkresiduals()` |

> Time series analysis requires understanding the data's structure — always plot first, decompose, check stationarity, then model.
