---
title: "Statistical Graphics"
subtitle: "Visualising Data Effectively in R"
author: "Pawan"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    highlight-style: github
  pdf:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

**Statistical graphics** are visual representations of data that help us explore distributions, relationships, patterns, and anomalies that are impossible to see in tables of numbers.

> "A picture is worth a thousand words — in statistics, a good chart is worth a thousand rows of data."

Good graphics are the first and last step in data analysis: used in **exploratory data analysis (EDA)** to understand data, and in **reporting** to communicate findings.

---

# Principles of Good Statistical Graphics

1. **Show the data**: Do not hide information with over-aggregation.
2. **Maximise data-ink ratio**: Remove non-essential decorations (chart junk).
3. **Label clearly**: Every axis, title, and legend must be informative.
4. **Choose the right chart**: The chart type must match the data type and the question.
5. **Avoid misleading scales**: Always start y-axis at 0 for bar charts.

---

# Choosing the Right Chart

| Data Type | Comparison Goal | Best Chart |
|-----------|----------------|-----------|
| One continuous variable | Distribution | Histogram, Density, Boxplot |
| One categorical variable | Frequency | Bar chart, Pie chart |
| Two continuous variables | Relationship | Scatter plot |
| Continuous over time | Trend | Line chart |
| Continuous by category | Comparison | Side-by-side boxplot |
| Multivariate | Overview | Pairs plot, Heatmap |
| Part-to-whole | Proportion | Stacked bar, Pie |

---

# Base R Graphics

## Setup

```{r}
data(mtcars)
data(iris)

# Convert cyl to factor for grouped plots
mtcars$cyl <- as.factor(mtcars$cyl)
```

## Histogram — Distribution of One Variable

```{r}
#| fig-width: 7
#| fig-height: 5

hist(mtcars$mpg,
     breaks = 12,
     col    = "steelblue",
     border = "white",
     main   = "Distribution of Fuel Efficiency",
     xlab   = "Miles Per Gallon (MPG)",
     ylab   = "Frequency")

# Add density curve
lines(density(mtcars$mpg) |> (\(d) list(x = d$x, y = d$y * length(mtcars$mpg) * diff(range(mtcars$mpg)) / 12))(),
      col = "red", lwd = 2)
```

**When to use**: To see the shape, spread, and central tendency of a single continuous variable.

## Boxplot — Comparing Distributions

```{r}
#| fig-width: 7
#| fig-height: 5

boxplot(mpg ~ cyl,
        data   = mtcars,
        col    = c("lightblue", "lightgreen", "salmon"),
        border = "navy",
        main   = "MPG by Number of Cylinders",
        xlab   = "Cylinders",
        ylab   = "Miles Per Gallon")
```

**Reading a boxplot:**

- **Box**: Q1 to Q3 (middle 50% of data)
- **Line inside box**: Median (Q2)
- **Whiskers**: Extend to 1.5 × IQR from the box edges
- **Points beyond whiskers**: Outliers

**When to use**: Compare distributions across groups; excellent for detecting outliers.

## Scatter Plot — Relationship Between Two Variables

```{r}
#| fig-width: 7
#| fig-height: 5

plot(mtcars$wt, mtcars$mpg,
     pch  = 16,
     col  = as.numeric(mtcars$cyl) + 1,
     cex  = 1.3,
     xlab = "Weight (1000 lbs)",
     ylab = "Miles Per Gallon",
     main = "Weight vs MPG by Cylinders")

legend("topright",
       legend = c("4 cyl", "6 cyl", "8 cyl"),
       col    = c(2, 3, 4),
       pch    = 16,
       title  = "Cylinders")

# Add overall trend line
abline(lm(mpg ~ wt, data = mtcars), col = "darkgrey", lwd = 2, lty = 2)
```

**When to use**: Examine the relationship (direction, strength, linearity) between two continuous variables.

## Bar Chart — Frequency of Categories

```{r}
#| fig-width: 7
#| fig-height: 5

cyl_counts <- table(mtcars$cyl)

barplot(cyl_counts,
        col    = c("lightblue", "steelblue", "navy"),
        border = "white",
        main   = "Number of Cars by Cylinder Count",
        xlab   = "Cylinders",
        ylab   = "Count",
        ylim   = c(0, 16))
```

**When to use**: Show counts or frequencies for categorical data.

## Line Chart — Trends Over Time

```{r}
#| fig-width: 9
#| fig-height: 4

plot(AirPassengers,
     col  = "steelblue",
     lwd  = 2,
     main = "Monthly Airline Passengers (1949–1960)",
     xlab = "Year",
     ylab = "Passengers (thousands)")
```

**When to use**: Display how a quantity changes over time.

## Pairs Plot — Multivariate Overview

```{r}
#| fig-width: 8
#| fig-height: 8

pairs(iris[, 1:4],
      col  = as.numeric(iris$Species) + 1,
      pch  = 16,
      main = "Pairs Plot — Iris Dataset",
      labels = c("Sepal\nLength", "Sepal\nWidth", "Petal\nLength", "Petal\nWidth"))
```

**When to use**: Explore all pairwise relationships in a dataset simultaneously. Excellent for EDA.

---

# ggplot2 — Grammar of Graphics

`ggplot2` is based on the **Grammar of Graphics**: every plot is built from:

1. **Data**: the dataset
2. **Aesthetics** (`aes`): what variables map to x, y, colour, size, shape
3. **Geometries** (`geom_*`): the visual elements (points, lines, bars, boxes)
4. **Scales**: control axes and legends
5. **Facets**: split into subplots
6. **Theme**: control the non-data appearance

```r
ggplot(data, aes(x = ..., y = ..., colour = ...)) +
  geom_*() +
  labs(...) +
  theme_*()
```

## Setup

```{r}
# install.packages("ggplot2")
library(ggplot2)

# Convert for ggplot
mtcars2 <- mtcars
mtcars2$cyl <- as.factor(mtcars2$cyl)
```

## ggplot2 Histogram

```{r}
#| fig-width: 7
#| fig-height: 4

ggplot(mtcars2, aes(x = mpg)) +
  geom_histogram(bins = 12, fill = "steelblue", colour = "white") +
  geom_density(aes(y = after_stat(count) * 2.5),
               colour = "red", linewidth = 1) +
  labs(title = "Distribution of MPG",
       x = "Miles Per Gallon",
       y = "Count") +
  theme_minimal()
```

## ggplot2 Boxplot

```{r}
#| fig-width: 7
#| fig-height: 4

ggplot(mtcars2, aes(x = cyl, y = mpg, fill = cyl)) +
  geom_boxplot(outlier.colour = "red", outlier.shape = 16) +
  scale_fill_manual(values = c("lightblue", "lightgreen", "salmon")) +
  labs(title = "MPG Distribution by Cylinders",
       x = "Number of Cylinders",
       y = "Miles Per Gallon",
       fill = "Cylinders") +
  theme_minimal() +
  theme(legend.position = "none")
```

## ggplot2 Scatter Plot with Smooth Trend

```{r}
#| fig-width: 7
#| fig-height: 4

ggplot(mtcars2, aes(x = wt, y = mpg, colour = cyl)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "black",
              linetype = "dashed", linewidth = 0.8) +
  scale_colour_manual(values = c("4" = "#2196F3",
                                  "6" = "#4CAF50",
                                  "8" = "#F44336")) +
  labs(title    = "Weight vs MPG",
       subtitle = "Linear trend with 95% confidence band",
       x        = "Weight (1000 lbs)",
       y        = "Miles Per Gallon",
       colour   = "Cylinders") +
  theme_minimal()
```

## ggplot2 Faceted Plot

```{r}
#| fig-width: 9
#| fig-height: 4

ggplot(mtcars2, aes(x = wt, y = mpg)) +
  geom_point(colour = "steelblue", size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, colour = "red") +
  facet_wrap(~ cyl, labeller = label_both) +
  labs(title = "MPG vs Weight — Faceted by Cylinders",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon") +
  theme_bw()
```

**When to use facets**: Compare the same relationship across subgroups without overlapping data.

## ggplot2 Bar Chart

```{r}
#| fig-width: 7
#| fig-height: 4

ggplot(mtcars2, aes(x = cyl, fill = cyl)) +
  geom_bar() +
  scale_fill_manual(values = c("lightblue", "steelblue", "navy")) +
  labs(title = "Car Count by Cylinders",
       x = "Cylinders", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Heatmap — Correlation Matrix

```{r}
#| fig-width: 7
#| fig-height: 6

# Compute correlation matrix
mtcars_num <- mtcars[, sapply(mtcars, is.numeric)]
cor_mat <- cor(mtcars_num)

# Reshape for ggplot
cor_df <- as.data.frame(as.table(cor_mat))
names(cor_df) <- c("Var1", "Var2", "Correlation")

ggplot(cor_df, aes(x = Var1, y = Var2, fill = Correlation)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = round(Correlation, 2)),
            size = 3, colour = "black") +
  scale_fill_gradient2(low = "#D73027", mid = "white", high = "#1A9850",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Heatmap — mtcars",
       x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**When to use**: Quickly identify strongly correlated variable pairs before regression modelling.

---

# Saving Plots

```{r}
#| eval: false

# Base R
png("my_plot.png", width = 800, height = 600)
hist(mtcars$mpg)
dev.off()

# ggplot2
p <- ggplot(mtcars2, aes(x = mpg)) + geom_histogram(fill = "steelblue")
ggsave("my_ggplot.png", plot = p, width = 8, height = 5, dpi = 300)
```

---

# Summary

| Chart Type | Best For | Base R | ggplot2 |
|------------|---------|--------|---------|
| Histogram | Distribution | `hist()` | `geom_histogram()` |
| Boxplot | Comparing groups | `boxplot()` | `geom_boxplot()` |
| Scatter | Relationships | `plot()` | `geom_point()` |
| Bar chart | Frequencies | `barplot()` | `geom_bar()` |
| Line chart | Time trends | `plot()` with `type="l"` | `geom_line()` |
| Pairs plot | Multivariate EDA | `pairs()` | `GGally::ggpairs()` |
| Heatmap | Correlations | `heatmap()` | `geom_tile()` |
| Density | Smooth distribution | `density()` + `plot()` | `geom_density()` |

> Statistical graphics are not decoration — they are the primary tool for discovering structure in data. Always plot before modelling, and always plot your results.
