---
title: "India GST Collection Analysis"
subtitle: "Interactive Dashboard — Financial Years 2017–18 to 2024–25"
author: "GST Analytics"
date: "2026-04-13"
date-modified: last-modified

format:
  html:
    theme:
      - cosmo
      - custom.scss          # optional — drop if not using Sass overrides
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    number-sections: false
    smooth-scroll: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 6
    fig-align: center
    embed-resources: true    # single self-contained HTML file
    page-layout: full
    df-print: paged

execute:
  echo: true
  warning: false
  message: false
  cache: false

knitr:
  opts_chunk:
    comment: "#>"
---

```{r setup}
#| label: setup
#| include: false

## ── Libraries ────────────────────────────────────────────────
library(data.table)
library(ggplot2)
library(ggiraph)
library(scales)

## ── Custom theme ─────────────────────────────────────────────
theme_gst <- function(base_size = 13) {
  theme_minimal(base_size = base_size) %+replace%
    theme(
      plot.background  = element_rect(fill = "#0F1923", color = NA),
      panel.background = element_rect(fill = "#131F2E", color = NA),
      panel.grid.major = element_line(color = "#1E3048", linewidth = 0.4),
      panel.grid.minor = element_blank(),
      panel.border     = element_blank(),
      plot.title       = element_text(color = "#F0F4F8", size = base_size + 4,
                                      face = "bold", hjust = 0,
                                      margin = margin(b = 4)),
      plot.subtitle    = element_text(color = "#8FA8C8", size = base_size - 1,
                                      hjust = 0, margin = margin(b = 12)),
      plot.caption     = element_text(color = "#4A6584", size = 9, hjust = 1),
      axis.title       = element_text(color = "#8FA8C8", size = base_size - 1),
      axis.text        = element_text(color = "#7A9AB8", size = base_size - 2),
      axis.text.x      = element_text(angle = 45, hjust = 1),
      strip.text       = element_text(color = "#C8DCEF", size = base_size - 1,
                                      face = "bold"),
      strip.background = element_rect(fill = "#1A2D44", color = NA),
      legend.background = element_rect(fill = "#0F1923", color = NA),
      legend.key        = element_rect(fill = "#0F1923", color = NA),
      legend.text       = element_text(color = "#8FA8C8", size = base_size - 2),
      legend.title      = element_text(color = "#C8DCEF", size = base_size - 1,
                                       face = "bold"),
      legend.position   = "top",
      plot.margin       = margin(16, 16, 12, 16)
    )
}

gst_colors <- c(
  CGST  = "#3B9EE0",
  SGST  = "#2DBE8E",
  IGST  = "#F5A623",
  CESS  = "#E05C5C",
  TOTAL = "#9B6CF5"
)

## ── Shared girafe options ────────────────────────────────────
girafe_opts <- function(border_color = "#3B9EE0") {
  list(
    opts_hover(css = "opacity:1;stroke:white;stroke-width:1.5px;"),
    opts_hover_inv(css = "opacity:0.2;"),
    opts_tooltip(css = paste0(
      "background:#1A2D44;color:#F0F4F8;",
      "border:1px solid ", border_color, ";border-radius:6px;",
      "padding:8px 12px;font-size:13px;font-family:monospace;"
    )),
    opts_toolbar(saveaspng = TRUE),
    opts_sizing(rescale = TRUE)
  )
}

fmt_cr <- function(x) {
  paste0("₹", formatC(round(x / 1000, 1), format = "f",
                       digits = 1, big.mark = ","), " K Cr")
}
```

```{r load-data}
#| label: load-data
#| cache: true

gst_data <- fread("/home/pawan/tax/gst.csv")

num_cols <- c("CGST", "SGST", "IGST", "CESS", "TOTAL")
gst_data[, (num_cols) := lapply(.SD, function(x) as.numeric(gsub(",", "", x))),
         .SDcols = num_cols]

latest_year  <- gst_data[!is.na(TOTAL) & TOTAL > 0, max(Year)]
exclude_rows <- c("Domestic GST Collection- All India",
                  "Gross Revenue",
                  "IGST and CESS on Import - All India")
```

## Overview

This interactive report explores **India's Goods and Services Tax (GST)** collection data across states and financial years. All monetary values are expressed in **₹ Thousands Crores (K Cr)**.

> **Tip:** Hover over any chart element to see detailed figures. Use the toolbar icon to download charts as PNG.

---

## 1 · All India Domestic GST Trends {#sec-all-india}

Year-wise breakdown of Central (CGST), State (SGST), Integrated (IGST) and Compensation Cess (CESS) for the consolidated All India basket.

```{r plot1-all-india}
#| label: plot1-all-india
#| fig-cap: "Stacked bars — hover to inspect individual tax components."

all_india_long <- melt(
  gst_data[State == "Domestic GST Collection- All India"],
  id.vars       = c("State", "Year"),
  measure.vars  = c("CGST", "SGST", "IGST", "CESS"),
  variable.name = "Tax_Type",
  value.name    = "Amount"
)
all_india_long[, tooltip := paste0(
  "<b>", Tax_Type, "</b><br>Year: ", Year,
  "<br>Amount: ", fmt_cr(Amount)
)]

p1 <- ggplot(all_india_long,
             aes(x = Year, y = Amount / 1000,
                 fill = Tax_Type, group = Tax_Type)) +
  geom_bar_interactive(
    stat = "identity",
    aes(tooltip = tooltip, data_id = paste(Year, Tax_Type)),
    width = 0.7, alpha = 0.88
  ) +
  scale_fill_manual(values = gst_colors) +
  scale_y_continuous(labels = label_comma(suffix = " K")) +
  labs(title    = "All India Domestic GST Collection Trends",
       subtitle = "Year-wise breakdown by tax type  (₹ Thousands Crores)",
       x = "Financial Year", y = "Amount (₹ Thousands Crores)",
       fill = "Tax Type",
       caption = "Source: GST Council / Ministry of Finance") +
  theme_gst()

girafe(ggobj = p1, options = girafe_opts("#3B9EE0"))
```

---

## 2 · GST Collection by State {#sec-states}

Top 16 states ranked by cumulative GST, broken down by tax type.

```{r plot2-states}
#| label: plot2-states
#| fig-height: 10
#| fig-cap: "Stacked bars per state — hover to see component amounts."

states_dt   <- gst_data[!State %in% exclude_rows]
top16_names <- states_dt[TOTAL > 0,
                          .(S = sum(TOTAL, na.rm = TRUE)), by = State
                         ][order(-S)][1:16, State]

states_long <- melt(
  states_dt[State %in% top16_names],
  id.vars       = c("State", "Year", "TOTAL"),
  measure.vars  = c("CGST", "SGST", "IGST", "CESS"),
  variable.name = "Tax_Type",
  value.name    = "Amount"
)
states_long[, tooltip := paste0(
  "<b>", State, "</b><br>Year: ", Year,
  "<br><i>", Tax_Type, "</i>: ", fmt_cr(Amount),
  "<br>Total: ", fmt_cr(TOTAL)
)]

p2 <- ggplot(states_long,
             aes(x = Year, y = Amount / 1000, fill = Tax_Type)) +
  geom_bar_interactive(
    stat = "identity",
    aes(tooltip = tooltip, data_id = paste(State, Year, Tax_Type)),
    alpha = 0.88
  ) +
  scale_fill_manual(values = gst_colors) +
  scale_y_continuous(labels = label_comma()) +
  facet_wrap(~State, scales = "free_y", ncol = 4) +
  labs(title    = "GST Collection by State",
       subtitle = "Top 16 states — stacked by tax type  (₹ Thousands Crores)",
       x = "Financial Year", y = "Amount (₹ Thousands Crores)",
       fill = "Tax Type", caption = "Source: GST Council") +
  theme_gst() +
  theme(axis.text.x = element_text(size = 7))

girafe(ggobj = p2, width_svg = 14, height_svg = 10,
       options = girafe_opts("#2DBE8E"))
```

---

## 3 · Major States — Collection Trends {#sec-major}

Line + point chart for 10 major states. Hover a point to see YoY growth.

```{r plot3-major-states}
#| label: plot3-major-states
#| fig-height: 7
#| fig-cap: "Each panel is one state; hover points reveal YoY growth %."

major_states <- c("Maharashtra", "Gujarat", "Tamil Nadu", "Karnataka",
                  "Uttar Pradesh", "West Bengal", "Delhi",
                  "Rajasthan", "Madhya Pradesh", "Telangana")

state_palette <- c(
  Maharashtra = "#3B9EE0", Gujarat = "#2DBE8E",
  `Tamil Nadu` = "#F5A623", Karnataka = "#E05C5C",
  `Uttar Pradesh` = "#9B6CF5", `West Bengal` = "#F0C060",
  Delhi = "#60C0F0", Rajasthan = "#E08050",
  `Madhya Pradesh` = "#70D080", Telangana = "#C070D0"
)

major_dt <- gst_data[State %in% major_states & !is.na(TOTAL) & TOTAL > 0
                     ][order(State, Year)]
major_dt[, Growth := round((TOTAL - shift(TOTAL)) / shift(TOTAL) * 100, 1),
         by = State]
major_dt[, tooltip := paste0(
  "<b>", State, "</b><br>Year: ", Year,
  "<br>Total: ", fmt_cr(TOTAL),
  ifelse(!is.na(Growth), paste0("<br>YoY: <b>", Growth, "%</b>"), "")
)]

p3 <- ggplot(major_dt,
             aes(x = Year, y = TOTAL / 1000,
                 color = State, group = State)) +
  geom_line_interactive(linewidth = 1.3,
                        aes(tooltip = State, data_id = State)) +
  geom_point_interactive(size = 2.5,
                         aes(tooltip = tooltip,
                             data_id = paste(State, Year))) +
  scale_color_manual(values = state_palette) +
  scale_y_continuous(labels = label_comma()) +
  facet_wrap(~State, ncol = 5, scales = "free_y") +
  labs(title    = "GST Collection Trends — Major States",
       subtitle = "Hover a point to see YoY growth  (₹ Thousands Crores)",
       x = "Financial Year", y = "Total GST (₹ Thousands Crores)",
       caption  = "Source: GST Council") +
  theme_gst() +
  theme(legend.position = "none")

girafe(ggobj = p3, width_svg = 14, height_svg = 7,
       options = girafe_opts("#F5A623"))
```

---

## 4 · GST Composition — `r latest_year` {#sec-composition}

Donut showing the share of each tax head for the latest available financial year across all states.

```{r plot4-composition}
#| label: plot4-composition
#| fig-height: 5
#| fig-cap: "Hover a segment to see the exact amount and share."

latest_comp <- gst_data[!State %in% exclude_rows & Year == latest_year,
                          .(CGST = sum(CGST, na.rm = TRUE),
                            SGST = sum(SGST, na.rm = TRUE),
                            IGST = sum(IGST, na.rm = TRUE),
                            CESS = sum(CESS, na.rm = TRUE))]

comp_long <- melt(latest_comp,
                  measure.vars  = c("CGST", "SGST", "IGST", "CESS"),
                  variable.name = "Tax_Type",
                  value.name    = "Amount")
comp_long[, Share := round(Amount / sum(Amount) * 100, 1)]
comp_long[, tooltip := paste0(
  "<b>", Tax_Type, "</b><br>",
  fmt_cr(Amount), "<br>Share: <b>", Share, "%</b>"
)]

p4 <- ggplot(comp_long,
             aes(x = "", y = Amount / 1000, fill = Tax_Type)) +
  geom_bar_interactive(
    stat = "identity", width = 1,
    aes(tooltip = tooltip, data_id = as.character(Tax_Type)),
    color = "#0F1923", linewidth = 1.2
  ) +
  coord_polar("y", start = 0) +
  geom_text(aes(label = paste0(Tax_Type, "\n", Share, "%")),
            position = position_stack(vjust = 0.55),
            color = "white", size = 4.2, fontface = "bold") +
  scale_fill_manual(values = gst_colors) +
  labs(title    = paste0("GST Composition — All States (", latest_year, ")"),
       subtitle = "Distribution across tax types",
       fill = "Tax Type", caption = "Source: GST Council") +
  theme_gst() +
  theme(axis.title = element_blank(), axis.text = element_blank(),
        panel.grid = element_blank())

girafe(ggobj = p4, options = girafe_opts("#9B6CF5"))
```

---

## 5 · Heatmap — State × Year {#sec-heatmap}

```{r plot5-heatmap}
#| label: plot5-heatmap
#| fig-height: 7
#| fig-cap: "Top 20 states by cumulative GST; brighter = higher collection."

heatmap_dt <- gst_data[
  !State %in% c(exclude_rows, "OIDAR", "Other Territory") & TOTAL > 0,
  .(State, Year, TOTAL)
]
top20 <- heatmap_dt[, .(S = sum(TOTAL, na.rm = TRUE)), by = State
                    ][order(-S)][1:20, State]

heatmap_dt <- heatmap_dt[State %in% top20]
heatmap_dt[, tooltip := paste0(
  "<b>", State, "</b><br>Year: ", Year, "<br>Total: ", fmt_cr(TOTAL)
)]

p5 <- ggplot(heatmap_dt,
             aes(x = Year,
                 y = reorder(State, TOTAL, FUN = sum),
                 fill = TOTAL / 1000)) +
  geom_tile_interactive(
    aes(tooltip = tooltip, data_id = paste(State, Year)),
    color = "#0F1923", linewidth = 0.4
  ) +
  scale_fill_gradient2(
    low = "#0D2137", mid = "#1E6FAD", high = "#FFD166",
    midpoint = median(heatmap_dt$TOTAL / 1000, na.rm = TRUE),
    name = "₹ K Crores", labels = label_comma()
  ) +
  labs(title    = "Heatmap — GST Collections by State & Year",
       subtitle = "Top 20 States/UTs  (₹ Thousands Crores)",
       x = "Financial Year", y = NULL,
       caption  = "Source: GST Council") +
  theme_gst() +
  theme(legend.position = "right",
        axis.text.y = element_text(size = 9))

girafe(ggobj = p5, width_svg = 12, height_svg = 7,
       options = girafe_opts("#FFD166"))
```

---

## 6 · YoY Growth Rate — All India {#sec-growth}

```{r plot6-growth}
#| label: plot6-growth
#| fig-cap: "Green = positive growth; red = contraction."

growth_dt <- gst_data[
  State == "Domestic GST Collection- All India" & !is.na(TOTAL) & TOTAL > 0
][order(Year)]
growth_dt[, Growth_Rate := round((TOTAL - shift(TOTAL)) / shift(TOTAL) * 100, 1)]
growth_dt[, tooltip := paste0(
  "<b>", Year, "</b><br>Growth: <b>",
  ifelse(!is.na(Growth_Rate), paste0(Growth_Rate, "%"), "N/A"),
  "</b><br>Total: ", fmt_cr(TOTAL)
)]
growth_dt[, bar_fill := ifelse(Growth_Rate >= 0, "Positive", "Negative")]

p6 <- ggplot(growth_dt[!is.na(Growth_Rate)],
             aes(x = Year, y = Growth_Rate, fill = bar_fill)) +
  geom_bar_interactive(
    stat = "identity", width = 0.65, alpha = 0.88,
    aes(tooltip = tooltip, data_id = Year)
  ) +
  geom_hline(yintercept = 0, color = "#8FA8C8", linewidth = 0.6) +
  geom_text(
    aes(label = paste0(Growth_Rate, "%"),
        vjust = ifelse(Growth_Rate >= 0, -0.4, 1.4)),
    color = "#C8DCEF", size = 3.5, fontface = "bold"
  ) +
  scale_fill_manual(values = c(Positive = "#2DBE8E", Negative = "#E05C5C"),
                    guide = "none") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(title    = "All India Domestic GST — YoY Growth Rate",
       subtitle = "Percentage change from previous financial year",
       x = "Financial Year", y = "Growth Rate (%)",
       caption  = "Source: GST Council") +
  theme_gst()

girafe(ggobj = p6, options = girafe_opts("#2DBE8E"))
```

---

## 7 · CGST vs SGST — Selected States {#sec-cgst-sgst}

```{r plot7-cgst-sgst}
#| label: plot7-cgst-sgst
#| fig-height: 7
#| fig-cap: "Dodged bars — CGST and SGST compared side by side."

comp_data  <- gst_data[State %in% c("Maharashtra", "Gujarat",
                                     "Tamil Nadu", "Karnataka") &
                          Year >= "2019-20"]
comp_long2 <- melt(comp_data,
                   id.vars       = c("State", "Year"),
                   measure.vars  = c("CGST", "SGST"),
                   variable.name = "Tax_Type",
                   value.name    = "Amount")
comp_long2[, tooltip := paste0(
  "<b>", State, "</b><br>Year: ", Year,
  "<br>", Tax_Type, ": ", fmt_cr(Amount)
)]

p7 <- ggplot(comp_long2,
             aes(x = Year, y = Amount / 1000, fill = Tax_Type)) +
  geom_bar_interactive(
    stat = "identity", position = "dodge", width = 0.7, alpha = 0.88,
    aes(tooltip = tooltip, data_id = paste(State, Year, Tax_Type))
  ) +
  scale_fill_manual(values = c(CGST = "#3B9EE0", SGST = "#2DBE8E")) +
  scale_y_continuous(labels = label_comma()) +
  facet_wrap(~State, scales = "free_y") +
  labs(title    = "CGST vs SGST — Major States",
       subtitle = "Central vs State GST since 2019-20  (₹ Thousands Crores)",
       x = "Financial Year", y = "Amount (₹ Thousands Crores)",
       fill = "Tax Type", caption = "Source: GST Council") +
  theme_gst()

girafe(ggobj = p7, width_svg = 11, height_svg = 7,
       options = girafe_opts("#3B9EE0"))
```

---

## 8 · CESS Contribution Trend {#sec-cess}

```{r plot8-cess}
#| label: plot8-cess
#| fig-cap: "Hover points to see exact CESS share per year."

cess_dt <- gst_data[!State %in% exclude_rows & TOTAL > 0,
                     .(CESS_Share = round(
                       sum(CESS, na.rm = TRUE) /
                         sum(TOTAL, na.rm = TRUE) * 100, 2)),
                     by = Year][order(Year)]
cess_dt[, tooltip := paste0(
  "<b>", Year, "</b><br>CESS Share: <b>", CESS_Share, "%</b>"
)]

p8 <- ggplot(cess_dt, aes(x = Year, y = CESS_Share, group = 1)) +
  geom_line_interactive(
    color = "#F5A623", linewidth = 1.4,
    aes(tooltip = "CESS trend", data_id = "line")
  ) +
  geom_point_interactive(
    color = "#E05C5C", size = 4, alpha = 0.9,
    aes(tooltip = tooltip, data_id = Year)
  ) +
  geom_text(aes(label = paste0(CESS_Share, "%")),
            vjust = -1, color = "#C8DCEF", size = 3.5) +
  scale_y_continuous(
    labels = label_percent(scale = 1),
    limits = c(0, max(cess_dt$CESS_Share, na.rm = TRUE) * 1.3)
  ) +
  labs(title    = "CESS Contribution to Total GST",
       subtitle = "Percentage share of CESS across all states, by year",
       x = "Financial Year", y = "CESS Share (%)",
       caption  = "Source: GST Council") +
  theme_gst()

girafe(ggobj = p8, options = girafe_opts("#F5A623"))
```

---

## 9 · Statistical Summary Tables {#sec-tables}

### Top 15 States by Average GST Collection

```{r table-avg}
#| label: table-avg

state_avg <- gst_data[
  !State %in% c(exclude_rows, "OIDAR", "Other Territory") & TOTAL > 0,
  .(
    `Avg (₹ K Cr)` = round(mean(TOTAL, na.rm = TRUE) / 1000, 2),
    `Max (₹ K Cr)` = round(max(TOTAL,  na.rm = TRUE) / 1000, 2),
    `Min (₹ K Cr)` = round(min(TOTAL,  na.rm = TRUE) / 1000, 2)
  ),
  by = State
][order(-`Avg (₹ K Cr)`)][1:15]

knitr::kable(state_avg, format.args = list(big.mark = ","))
```

### Top 10 Growth Leaders (2023-24 → 2024-25)

```{r table-growth}
#| label: table-growth

growth_by_state <- gst_data[
  !State %in% c(exclude_rows, "OIDAR", "Other Territory") &
    Year %in% c("2023-24", "2024-25") & TOTAL > 0
]
growth_wide <- dcast(growth_by_state, State ~ Year, value.var = "TOTAL")
growth_wide[, `Growth (%)` := round((`2024-25` - `2023-24`) / `2023-24` * 100, 2)]

knitr::kable(
  growth_wide[order(-`Growth (%)`)][1:10,
    .(State, `2023-24 (₹ Cr)` = `2023-24`,
             `2024-25 (₹ Cr)` = `2024-25`,
             `Growth (%)`)],
  format.args = list(big.mark = ",")
)
```

---

## Session Info {#sec-session}

```{r session}
#| label: session
#| code-fold: true
#| code-summary: "R session info"
sessionInfo()
```
