---
title: "India's Public Foodgrain Procurement: Trends, Patterns & Policy Implications"
description: "A detailed analysis of foodgrain procurement by FCI and state agencies across marketing seasons, with regional breakdowns and MSP evolution."
author: "Pawan"
date: "2025/10/30"
date-modified: last-modified
categories:
  - "Food Economy"
  - "Procurement"
  - "Agriculture Policy"
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
    theme: cosmo
editor_options:
  chunk_output_type: console
execute:
  error: false
  message: false
  warning: false
  eval: true
---

## Overview

India operates one of the world's largest public food procurement systems. The government procures wheat (rabi), paddy/rice (kharif), and coarse grains at the **Minimum Support Price (MSP)** through the Food Corporation of India (FCI) and state government agencies. Procured stocks underpin the **Public Distribution System (PDS)**, midday-meal schemes, and the National Food Security Act (NFSA) welfare programmes.

**Key policy features:**
- Procurement is price-floor insurance for farmers — FCI and state agencies buy all offered quantities at MSP.
- Stocks feed welfare: NFSA covers 80 crore beneficiaries with 5 kg/person/month of subsidised grain.
- Buffer norms guide minimum stock requirements for food security.

---

## Setup

```{r setup}
library(data.table)
library(ggplot2)
library(ggiraph)
library(scales)
library(gt)
library(plotly)
library(viridis)
library(patchwork)

my_theme <- theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(size = 16, face = "bold", hjust = 0.5, margin = margin(b = 8)),
    plot.subtitle   = element_text(size = 11, hjust = 0.5, color = "#7F8C8D"),
    plot.caption    = element_text(size = 9, color = "#95A5A6", hjust = 0),
    axis.title      = element_text(size = 12, face = "bold"),
    axis.text       = element_text(size = 10),
    axis.text.x     = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position = "bottom",
    legend.title    = element_blank(),
    panel.grid.major= element_line(color = "#ECF0F1", linewidth = 0.4),
    panel.grid.minor= element_blank(),
    strip.background= element_rect(fill = "#F8F9FA", color = NA),
    strip.text      = element_text(face = "bold", size = 11)
  )
```

---

## Section 1 · Total Procurement by Crop

```{r procurement-total}
t1 <- fread("/home/pawan/procurement/procurement.csv")

## Summarise by crop and marketing season
proc_sum <- t1[, .(Total     = sum(Total, na.rm=TRUE),
                   FCI       = sum(FCI, na.rm=TRUE),
                   StateAgency = sum(StateAgency, na.rm=TRUE)),
               by = .(Crop, MarketingSeason)]

proc_sum[, FCI_Share   := round(FCI/Total*100, 1)]
proc_sum[, State_Share := round(StateAgency/Total*100, 1)]
proc_sum[, tt := paste0("<b>",Crop,"</b> | Season: ",MarketingSeason,
                         "<br>Total: ",round(Total,1)," LMT",
                         "<br>FCI: ",round(FCI,1)," (",FCI_Share,"%)",
                         "<br>States: ",round(StateAgency,1)," (",State_Share,"%)")]

## By-crop faceted bar
p1 <- ggplot(proc_sum, aes(x=MarketingSeason, y=Total, fill=Crop)) +
  geom_bar_interactive(aes(tooltip=tt, data_id=paste(Crop,MarketingSeason)),
                        stat="identity", width=0.75, alpha=0.9) +
  facet_wrap(~Crop, scales="free_y") +
  scale_fill_brewer(palette="Set2") +
  scale_y_continuous(labels=comma_format()) +
  my_theme + theme(legend.position="none") +
  labs(title="Foodgrain Procurement by Crop & Marketing Season",
       subtitle="Hover over bars for FCI vs state agency breakdown",
       y="Quantity (Lakh Metric Tonnes)", x="Marketing Season",
       caption="Source: DFPD Procurement Portal | Analysis: Pawan")

girafe(ggobj=p1, width_svg=13, height_svg=7,
       options=list(
         opts_tooltip(css="background:#fff;color:#333;padding:10px;border-radius:6px;border:1px solid #ddd;font-size:12px;"),
         opts_hover(css="opacity:0.9;stroke:#333;stroke-width:1.5;"),
         opts_hover_inv(css="opacity:0.3;"),
         opts_zoom(min=0.8, max=3),
         opts_sizing(rescale=TRUE, width=1)))
```

---

## Section 2 · FCI vs State Agency Split

The procurement machinery involves two tiers: FCI (central agency) and state government agencies. Their relative roles have shifted over time — states have taken on more responsibility as part of decentralisation.

```{r fci-state-split}
proc_long <- melt(proc_sum[, .(Crop,MarketingSeason,FCI,StateAgency)],
                  id.vars=c("Crop","MarketingSeason"),
                  variable.name="Agency", value.name="Qty")
proc_long[, tt := paste0(Agency,"\nCrop: ",Crop,"\nSeason: ",MarketingSeason,
                          "\n",round(Qty,1)," LMT")]

p2 <- ggplot(proc_long, aes(x=MarketingSeason, y=Qty, fill=Agency)) +
  geom_bar_interactive(aes(tooltip=tt, data_id=Agency),
                        stat="identity", position="stack", width=0.75, alpha=0.9) +
  facet_wrap(~Crop, scales="free_y") +
  scale_fill_manual(values=c("FCI"="#2C3E50","StateAgency"="#E67E22"),
                    labels=c("FCI"="FCI","StateAgency"="State Agencies")) +
  scale_y_continuous(labels=comma_format()) +
  my_theme +
  labs(title="FCI vs State Agency Procurement Split",
       subtitle="States increasingly driving procurement in newer seasons",
       y="Quantity (LMT)", x="Marketing Season",
       caption="Source: DFPD | Analysis: Pawan")

girafe(ggobj=p2, width_svg=13, height_svg=7,
       options=list(
         opts_tooltip(css="background:#fff;color:#333;padding:8px;border-radius:5px;border:1px solid #ddd;"),
         opts_hover(css="opacity:0.85;stroke-width:1.5;"),
         opts_sizing(rescale=TRUE, width=1)))
```

---

## Section 3 · Procurement Trend Analysis

```{r trend-analysis}
## Total annual procurement across all crops
annual_total <- proc_sum[, .(Total_All = sum(Total)), by=MarketingSeason]
setorder(annual_total, MarketingSeason)
annual_total[, YoY := round((Total_All/shift(Total_All)-1)*100, 1)]

## Line chart with growth annotations
p3 <- ggplot(annual_total, aes(x=MarketingSeason, y=Total_All, group=1,
                                text=paste0(MarketingSeason,"\n",round(Total_All,1)," LMT",
                                            "\nYoY: ",YoY,"%"))) +
  geom_col(fill="#3498DB", alpha=0.4, width=0.7) +
  geom_line(color="#2C3E50", linewidth=1.3) +
  geom_point(color="#2C3E50", size=3.5) +
  geom_text(aes(label=ifelse(!is.na(YoY), paste0(YoY,"%"), "")),
            vjust=-1.2, size=3.2, color="#E74C3C") +
  my_theme +
  labs(title="Total Foodgrain Procurement — All Crops",
       subtitle="YoY growth rates annotated in red",
       y="Quantity (LMT)", x="Marketing Season",
       caption="Source: DFPD | Analysis: Pawan")

ggplotly(p3, tooltip="text") |> layout(hovermode="x unified")
```

---

## Section 4 · Summary Statistics Table

```{r summary-table}
proc_stats <- proc_sum[, .(
  Avg_Procurement = round(mean(Total),1),
  Max_Procurement = round(max(Total),1),
  Min_Procurement = round(min(Total),1),
  Avg_FCI_Share   = round(mean(FCI_Share),1),
  Seasons         = .N
), by=Crop]

proc_stats |>
  gt() |>
  tab_header(title="Procurement Summary Statistics by Crop") |>
  cols_label(Crop="Crop", Avg_Procurement="Avg (LMT)",
             Max_Procurement="Max (LMT)", Min_Procurement="Min (LMT)",
             Avg_FCI_Share="Avg FCI Share %", Seasons="No. of Seasons") |>
  fmt_number(columns=2:4, decimals=1) |>
  data_color(Avg_Procurement,
             colors=scales::col_numeric(c("#D5F5E3","#1A5276"), domain=NULL)) |>
  opt_stylize(style=3) |>
  tab_source_note("Source: DFPD Central Food Grains Procurement Portal")
```

---

## Section 5 · Policy Context

```{r policy-notes}
## Paddy procurement regional breakdown (if available)
if ("State" %in% names(t1)) {
  paddy_state <- t1[Crop=="Paddy",
    .(Total=sum(Total,na.rm=TRUE)), .(State, MarketingSeason)]
  top_states <- paddy_state[, .(Total=sum(Total)), by=State][order(-Total)][1:10, State]
  paddy_top  <- paddy_state[State %in% top_states]

  p_state <- ggplot(paddy_top, aes(x=MarketingSeason, y=Total, fill=State)) +
    geom_bar(stat="identity", position="stack", width=0.75, alpha=0.9) +
    scale_fill_viridis_d() +
    my_theme +
    labs(title="Paddy Procurement by Top States",
         y="Quantity (LMT)", x="Marketing Season",
         caption="Source: DFPD | Analysis: Pawan")
  ggplotly(p_state) |> layout(hovermode="x unified")
}
```

> **Key insights:**
>
> 1. **Paddy dominates** procurement in volume terms, with Chhattisgarh, MP, Punjab and Haryana as the largest procuring states.
> 2. **Wheat procurement** fluctuates more sharply due to production variability in the Indo-Gangetic plain.
> 3. **State agencies** have grown as a share of total procurement, reflecting greater state-level operational capacity.
> 4. **COVID-19 impact** (2020–21): procurement held firm, underpinning the free-grain scheme (PMGKAY) for 80 crore beneficiaries.
