---
title: "India's Foodgrain Buffer Stocks: Trends, Norms, and Policy"
description: "Interactive analysis of India's central pool foodgrain stocks — rice, wheat, unmilled paddy, and coarse grains — from 1972 to 2025, tracking stock positions against buffer norms and examining recent seasonal patterns."
author: "Pawan"
date: "2025/11/01"
date-modified: last-modified
categories:
  - "Food Economy"
  - "Agriculture"
  - "Public Policy"
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
editor_options:
  chunk_output_type: console
execute:
  error: false
  message: false
  warning: false
  eval: true
---

## Overview {.unnumbered}

India maintains one of the world's largest publicly managed foodgrain buffer stock systems. The **Food Corporation of India (FCI)** and state agencies procure, store, and distribute wheat and rice to approximately 800 million beneficiaries under the National Food Security Act (NFSA). The central pool stock position — reported monthly by FCI — is a critical indicator of India's food security preparedness, the government's procurement performance, and the adequacy of storage infrastructure.

---

::: {.callout-tip title="Key Findings at a Glance"}
- 📦 **Total foodgrain stocks peaked at over 100 LMT (10 MT)** in early 2025, driven by large unmilled paddy holdings
- 🌾 **Wheat stocks show sharp seasonal cycles**: declining through the year as distribution continues, then replenishing sharply after the April–June procurement season
- 🍚 **Rice stocks have grown steadily** from ~21 LMT (early 2023) to ~34 LMT (early 2025)
- ⚠️ **Wheat stocks hit a near-decade low of ~7.5 LMT** in April 2024, raising brief food security concerns before the 2024 rabi harvest
- 📊 **Stocks consistently exceed buffer norms**: India over-procures relative to minimum requirements, creating storage pressure and fiscal costs
- 🌾 **Unmilled paddy (dhaan) holdings have grown** as procurement has expanded from traditional rice-bowl states to eastern India
:::

---

## Setup

```{r setup}
library(data.table)
library(ggplot2)
library(ggiraph)
library(tidyr)
library(scales)
library(lubridate)
library(DT)

my_theme <- theme_minimal(base_size = 13) +
  theme(
    text             = element_text(color = "#2c3e50"),
    plot.title       = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle    = element_text(size = 11, hjust = 0.5, color = "#7f8c8d"),
    plot.caption     = element_text(size = 9, color = "#95a5a6", hjust = 0),
    axis.title       = element_text(size = 12, face = "bold"),
    axis.text        = element_text(size = 10),
    axis.text.x      = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position  = "bottom",
    legend.title     = element_blank(),
    panel.grid.major = element_line(color = "#ecf0f1", linewidth = 0.5),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#f8f9fa", color = NA),
    strip.text       = element_text(face = "bold")
  )

std_opts <- list(
  opts_tooltip(
    css = paste0("background-color:#2c3e50;color:white;padding:8px 12px;",
                 "border-radius:6px;font-size:12px;line-height:1.5;")
  ),
  opts_hover(css = "fill-opacity:0.85;stroke:white;stroke-width:1;"),
  opts_hover_inv(css = "opacity:0.2;"),
  opts_zoom(min = 0.5, max = 3),
  opts_toolbar(position = "topright"),
  opts_sizing(rescale = TRUE, width = 1)
)
```

## Data Loading and Preparation

```{r data}
## Load current stocks data
stocksall <- fread("~/stocks/stocksall.csv", sep = ",", header = TRUE)
stocksall <- unique(stocksall)

## Load historical data
oldstocks <- fread("~/stocks/pdsold.csv", sep = ",", header = TRUE)
oldstocks[, coarsegrains := stocks_total - stocks_rice - stocks_wheat]

## Reshape current stocks to long format
stocks_long <- melt(stocksall, id = c("Year", "Commodity"))
stocks_long[, Date := as.Date(paste("01", variable, Year, sep = "-"), "%d-%b-%Y")]
stocks_long[, Commodity := factor(Commodity,
  levels = c("Rice", "Wheat", "Unmilled Paddy", "Coarsegrain", "Total"))]

## Process historical data
oldstocks_clean <- separate(oldstocks, col = Year, sep = "-", into = c("v1", "year"))
setDT(oldstocks_clean)
oldstocks_clean[, Date := as.Date(paste("01", "04", year, sep = "-"), "%d-%m-%y")]
oldstocks_clean <- oldstocks_clean[Date < as.Date("2003-01-01"),
  .(Date, Rice = stocks_rice, Wheat = stocks_wheat, Coarsegrain = coarsegrains)]
old_long <- melt(oldstocks_clean, id = "Date", variable.name = "Commodity")
old_long[, value := value * 10]   ## → lakh tonnes
old_long[, Commodity := as.character(Commodity)]

## Filter and combine
new_long <- stocks_long[Commodity != "Total", .(Date, Commodity = as.character(Commodity), value)]
new_long <- new_long[!is.na(value)]
old_long <- old_long[!is.na(value)]

combined <- rbind(new_long, old_long)
combined[, value := as.numeric(value) / 10]   ## → million tonnes
combined <- combined[!is.na(value)]
combined[, Commodity := factor(Commodity,
  levels = rev(c("Rice", "Wheat", "Unmilled Paddy", "Coarsegrain")))]

## Build tooltip
combined[, tip := paste0("<b>", Commodity, "</b><br>",
                         "Date: ", format(Date, "%b %Y"), "<br>",
                         "Stock: <b>", round(value, 2), " MT</b>")]
```

---

## Full Historical Stock Position (1972–2025)

```{r full-history}
p_full <- ggplot(combined, aes(x = Date, y = value, fill = Commodity,
                                tooltip = tip, data_id = paste(Date, Commodity),
                                group = Commodity)) +
  geom_area_interactive(alpha = 0.85, color = "white", linewidth = 0.3,
                        position = "stack") +
  scale_fill_manual(values = c(
    "Rice"          = "#4E79A7",
    "Wheat"         = "#F28E2B",
    "Unmilled Paddy" = "#E15759",
    "Coarsegrain"   = "#76B7B2"
  )) +
  ## Mark major policy events
  annotate("rect", xmin = as.Date("2001-01-01"), xmax = as.Date("2003-01-01"),
           ymin = -Inf, ymax = Inf, fill = "#e74c3c", alpha = 0.07) +
  annotate("text", x = as.Date("2002-01-01"), y = 75,
           label = "2002\nDrought", size = 3, color = "#c0392b", angle = 90) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_y_continuous("Stocks (Million Tonnes)", labels = comma_format(),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title   = "India's Central Pool Foodgrain Stock Position (1972–2025)",
       subtitle = "Stacked area — hover for commodity breakdown by date",
       x = "Year",
       caption = "Source: Food Corporation of India (FCI) | Analysis: Pawan") +
  my_theme

girafe(ggobj = p_full, width_svg = 14, height_svg = 7, options = std_opts)
```

---

## Recent Trends: 2018–2025

```{r recent}
recent <- combined[Date >= as.Date("2018-01-01")]
recent[, tip_r := paste0("<b>", Commodity, "</b><br>",
                         "Date: ", format(Date, "%b %Y"), "<br>",
                         "Stock: <b>", round(value, 2), " MT</b>")]

p_recent <- ggplot(recent, aes(x = Date, y = value, fill = Commodity,
                                tooltip = tip_r, data_id = paste(Date, Commodity),
                                group = Commodity)) +
  geom_area_interactive(alpha = 0.85, color = "white", linewidth = 0.3,
                        position = "stack") +
  ## COVID lockdown marker
  geom_vline_interactive(aes(xintercept = as.Date("2020-03-25"),
                             tooltip = "National Lockdown — 25 March 2020"),
                         linetype = "dashed", color = "#e74c3c", linewidth = 0.7) +
  ## 2023 procurement year
  geom_vline_interactive(aes(xintercept = as.Date("2023-04-01"),
                             tooltip = "Start of KMS 2023–24"),
                         linetype = "dashed", color = "#2980b9", linewidth = 0.7) +
  annotate("text", x = as.Date("2020-06-01"), y = 70,
           label = "COVID\nLockdown", size = 3, color = "#e74c3c") +
  scale_fill_manual(values = c(
    "Rice"           = "#4E79A7",
    "Wheat"          = "#F28E2B",
    "Unmilled Paddy" = "#E15759",
    "Coarsegrain"    = "#76B7B2"
  )) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous("Stocks (Million Tonnes)", labels = comma_format(),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(title   = "Foodgrain Stocks: Recent Trends (2018–2025)",
       subtitle = "Seasonal wheat cycle and growing paddy/rice accumulation are clearly visible",
       x = "Month-Year",
       caption = "Source: FCI | Dashed lines mark COVID lockdown and KMS 2023–24 start | Analysis: Pawan") +
  my_theme + theme(axis.text.x = element_text(size = 8))

girafe(ggobj = p_recent, width_svg = 14, height_svg = 7, options = std_opts)
```

---

## Individual Commodity Trends

```{r individual}
p_ind <- ggplot(recent, aes(x = Date, y = value, color = Commodity,
                             tooltip = tip_r, data_id = paste(Date, Commodity),
                             group = Commodity)) +
  geom_line_interactive(linewidth = 1.2) +
  facet_wrap(~ Commodity, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Rice"           = "#4E79A7",
    "Wheat"          = "#F28E2B",
    "Unmilled Paddy" = "#E15759",
    "Coarsegrain"    = "#76B7B2"
  )) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = comma_format(suffix = " MT")) +
  labs(title   = "Individual Commodity Stock Trends (2018–2025)",
       x = "Date", y = "Stocks (Million Tonnes)",
       caption = "Source: FCI | Analysis: Pawan") +
  my_theme + theme(legend.position = "none")

girafe(ggobj = p_ind, width_svg = 12, height_svg = 8, options = std_opts)
```

---

## Last 24 Months — Interactive Data Table

```{r dt-table}
last24 <- combined[Date >= as.Date("2023-04-01")]
last24_wide <- dcast(last24, Date ~ Commodity, value.var = "value", fun.aggregate = sum)
last24_wide[, Total := rowSums(.SD, na.rm = TRUE), .SDcols = setdiff(names(last24_wide), "Date")]
last24_wide[, Date := format(Date, "%b-%Y")]

datatable(
  last24_wide,
  rownames = FALSE,
  extensions = "Buttons",
  options = list(
    pageLength = 12,
    dom = "Bfrtip",
    buttons = c("copy", "csv", "excel"),
    scrollX = TRUE
  ),
  caption = htmltools::tags$caption(
    style = "caption-side:top;text-align:center;font-size:15px;font-weight:bold;",
    "Central Pool Foodgrain Stocks — Last 24 Months (Million Tonnes)"
  )
) |>
  formatRound(columns = setdiff(names(last24_wide), "Date"), digits = 2)
```

---

## Key Takeaways

::: {.callout-important title="Summary of Findings"}
1. **India's central pool stocks are at historically high levels** relative to buffer norms, reflecting several consecutive good harvests and aggressive procurement — but also significant storage infrastructure pressure and fiscal carrying costs.

2. **Wheat stocks follow a pronounced seasonal cycle**: they decline through the year as PDS offtake and open market sales proceed, then jump sharply in April–June as the rabi wheat crop is procured. The April 2024 dip to ~7.5 LMT wheat was the lowest in nearly a decade — a reminder that even apparently comfortable stock positions can dissipate rapidly.

3. **Rice and unmilled paddy stocks have grown substantially**, driven by the expansion of paddy procurement from eastern states (Chhattisgarh, Telangana, Odisha, West Bengal) under the Decentralised Procurement scheme.

4. **Unmilled paddy holdings** represent a distinctive feature of India's system: paddy is often stored unmilled until conversion to rice is needed, creating a large but operationally more complex buffer than milled rice holdings.

5. **COVID-19 (2020) did not disrupt stocks** — the government used its large pre-pandemic stock position to mount the PM-GKAY free grain distribution scheme, demonstrating the food security value of the buffer stock system.

6. **The coarse grain stock position remains negligible** despite policy attention to millets — the procurement system for coarse grains outside a few states remains underdeveloped.
:::
