---
title: "India's Trade Statistics: Exports, Imports & Sectoral Composition"
description: "Interactive analysis of India's total merchandise trade — exports, imports, trade balance, and sector-level composition including cereals, fertilizers, and petroleum — using HS code classification."
author: "Pawan"
date: "2025/10/29"
date-modified: last-modified
categories:
  - "Trade"
  - "Economy"
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
    number-sections: true
    theme: flatly
editor_options:
  chunk_output_type: console
execute:
  error: false
  message: false
  warning: false
  eval: true
---

## Overview {.unnumbered}

India's merchandise trade has grown from under USD 200 billion (2005) to over USD 1.1 trillion (2023), making it the world's 10th largest merchandise trading nation. This analysis examines **total export and import trends**, the **trade balance trajectory**, and the composition of key sectors — cereals, fertilizers, petroleum, and edible oils — using Harmonised System (HS) code classification.

---

::: {.callout-tip title="Key Findings at a Glance"}
- 📈 **India's exports crossed USD 450 billion** in 2022–23, up from ~USD 170 billion a decade earlier
- 📉 **Trade deficit is structural and widening**: typically USD 200–250 billion annually in recent years
- 🌾 **Rice dominates cereal exports**: India is the world's largest rice exporter, accounting for ~40% of global trade
- ⛽ **Petroleum products are India's top export category** — a testament to refinery capacity expansions
- 🌱 **Fertilizer imports are large and import-intensive**: India imports ~$10 billion worth of fertilizers annually
- 🔄 **Export coverage ratio has declined** from ~70% (early 2000s) to ~55% (recent years), indicating growing import dependence
:::

---

## Setup

```{r setup}
library(data.table)
library(ggplot2)
library(ggiraph)
library(readxl)
library(viridis)
library(scales)

my_theme <- theme_minimal(base_size = 13) +
  theme(
    text             = element_text(color = "#2c3e50"),
    plot.title       = element_text(size = 16, face = "bold", hjust = 0.5,
                                    margin = margin(8, 0, 8, 0)),
    plot.subtitle    = element_text(size = 11, hjust = 0.5, color = "#7f8c8d"),
    plot.caption     = element_text(size = 9, color = "#95a5a6", hjust = 0),
    axis.title       = element_text(size = 12, face = "bold"),
    axis.text        = element_text(size = 10),
    axis.text.x      = element_text(angle = 45, vjust = 1, hjust = 1),
    legend.position  = "bottom",
    legend.title     = element_blank(),
    panel.grid.major = element_line(color = "#ecf0f1", linewidth = 0.5),
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "#f8f9fa", color = NA),
    strip.text       = element_text(face = "bold")
  )

std_opts <- list(
  opts_tooltip(
    css = paste0("background-color:#2c3e50;color:white;padding:8px 12px;",
                 "border-radius:6px;font-size:12px;line-height:1.5;")
  ),
  opts_hover(css = "stroke-width:2.5;opacity:1;"),
  opts_hover_inv(css = "opacity:0.2;"),
  opts_zoom(min = 0.5, max = 3),
  opts_toolbar(position = "topright"),
  opts_sizing(rescale = TRUE, width = 1)
)
```

## Data Loading & Cleaning

```{r data-load}
usd <- as.data.table(readRDS("/home/pawan/trade/totaltrade/total_exportsusd.rds"))
qty <- as.data.table(readRDS("/home/pawan/trade/totaltrade/total_exportsqty.rds"))
musd <- as.data.table(readRDS("/home/pawan/trade/totaltrade/total_importsusd.rds"))
mqty <- as.data.table(readRDS("/home/pawan/trade/totaltrade/total_importsqty.rds"))
twodigit <- as.data.table(read_excel("/home/pawan/trade/totaltrade/two-digit.xlsx"))

## Standardised cleaning function
clean_trade <- function(dt, val_col) {
  dt[, `:=`(HSCode2 = substr(HSCode, 1, 2),
            HSCode4 = substr(HSCode, 1, 4),
            HSCode6 = substr(HSCode, 1, 6))]
  dt[, (val_col) := as.numeric(gsub(",", "", get(val_col)))]
  dt[is.na(get(val_col)), (val_col) := 0]
  dt
}

usd  <- clean_trade(usd,  "Export")
qty  <- clean_trade(qty,  "Export")
musd <- clean_trade(musd, "Import")
mqty <- clean_trade(mqty, "Import")

## Merge HS classification
twodigit[, HSCode2 := as.numeric(HSCode2)]
qty[,  HSCode2 := as.numeric(HSCode2)]
mqty[, HSCode2 := as.numeric(HSCode2)]
qty  <- merge(qty,  twodigit, by = "HSCode2", all.x = TRUE)
mqty <- merge(mqty, twodigit, by = "HSCode2", all.x = TRUE)

## Aggregate totals
export_usd <- usd[,  .(Export_USD = sum(Export, na.rm = TRUE)), by = Year]
import_usd <- musd[, .(Import_USD = sum(Import, na.rm = TRUE)), by = Year]
export_qty <- qty[,  .(Export_Qty = sum(Export, na.rm = TRUE)), by = Year]
import_qty <- mqty[, .(Import_Qty = sum(Import, na.rm = TRUE)), by = Year]

trade <- Reduce(function(a, b) merge(a, b, by = "Year"),
                list(export_usd, import_usd, export_qty, import_qty))
trade[, `:=`(
  Trade_Balance   = Export_USD - Import_USD,
  Trade_Volume    = Export_USD + Import_USD,
  Export_Coverage = round(Export_USD * 100 / Import_USD, 1),
  Export_Growth   = round((Export_USD / shift(Export_USD) - 1) * 100, 1),
  Import_Growth   = round((Import_USD / shift(Import_USD) - 1) * 100, 1)
)]
```

---

## Trade Balance Trend

```{r trade-balance}
trade_long <- melt(trade[, .(Year, Export_USD, Import_USD)],
                   id = "Year", variable.name = "Flow", value.name = "USD")
trade_long[, tip := paste0("<b>", ifelse(Flow == "Export_USD", "Exports", "Imports"), "</b><br>",
                           "Year: ", Year, "<br>",
                           "Value: USD <b>", format(round(USD / 1000), big.mark = ","), "</b> Bn")]

pt <- ggplot() +
  ## Trade balance area
  geom_area(data = trade, aes(x = Year, y = Trade_Balance / 1000, group = 1),
            fill = "#e74c3c", alpha = 0.15) +
  ## Export & import lines
  geom_line_interactive(
    data = trade_long,
    aes(x = Year, y = USD / 1000, color = Flow, tooltip = tip,
        data_id = paste(Flow, Year), group = Flow),
    linewidth = 1.3
  ) +
  geom_point_interactive(
    data = trade_long,
    aes(x = Year, y = USD / 1000, color = Flow, tooltip = tip,
        data_id = paste(Flow, Year)),
    size = 2.5
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#7f8c8d") +
  scale_color_manual(values = c("Export_USD" = "#27ae60", "Import_USD" = "#e74c3c"),
                     labels = c("Export_USD" = "Exports", "Import_USD" = "Imports")) +
  scale_y_continuous(labels = label_dollar(suffix = "B", scale = 1)) +
  labs(title   = "India's Total Merchandise Trade",
       subtitle = "Exports, imports, and trade deficit (USD Billions)",
       x = "Year", y = "USD Billions",
       caption = "Source: Ministry of Commerce, India | Analysis: Pawan") +
  my_theme

girafe(ggobj = pt, width_svg = 11, height_svg = 6, options = std_opts)
```

### Export Coverage Ratio

```{r coverage-ratio}
trade[, tip_cov := paste0("Year: ", Year, "<br>",
                          "Export coverage: <b>", Export_Coverage, "%</b><br>",
                          "(Exports ÷ Imports × 100)")]
pc <- ggplot(trade[!is.na(Export_Coverage)],
             aes(x = Year, y = Export_Coverage, tooltip = tip_cov, data_id = Year, group = 1)) +
  geom_col_interactive(
    aes(fill = Export_Coverage < 60),
    width = 0.75, alpha = 0.85
  ) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "#27ae60", linewidth = 0.8) +
  geom_hline(yintercept = 60,  linetype = "dotted", color = "#e74c3c",  linewidth = 0.8) +
  scale_fill_manual(values = c("FALSE" = "#3498db", "TRUE" = "#e67e22"), guide = "none") +
  scale_y_continuous(labels = label_percent(scale = 1)) +
  labs(title   = "Export Coverage Ratio",
       subtitle = "How much of India's imports are covered by export earnings?",
       x = "Year", y = "Exports as % of Imports",
       caption = "Green dashed = 100% (balanced trade); Red dotted = 60% threshold") +
  my_theme

girafe(ggobj = pc, width_svg = 10, height_svg = 5, options = std_opts)
```

---

## Cereal Exports: Composition

India is the world's largest rice exporter and a significant wheat exporter in surplus years. The 2022–23 wheat export ban reflects India's policy of managing domestic food price inflation.

```{r cereals}
cereals <- qty[HSCode2 == 10]
cereals[, Commodity := fcase(
  HSCode4 == "1006", "Rice",
  HSCode4 == "1001", "Wheat",
  HSCode4 == "1005", "Maize",
  HSCode4 == "1007", "Grain Sorghum",
  HSCode4 == "1008", "Millet & Buckwheat",
  default = "Other Cereals"
)]
cereals_yr <- cereals[, .(Qty = sum(Export, na.rm = TRUE)), by = .(Year, Commodity)]
cereals_yr[, Total := sum(Qty), by = Year]
cereals_yr[, Share := round(Qty * 100 / Total)]
cereals_yr[, Commodity := factor(Commodity,
  levels = c("Other Cereals", "Millet & Buckwheat", "Grain Sorghum",
             "Maize", "Wheat", "Rice"))]
cereals_yr[, tip := paste0("<b>", Commodity, "</b><br>",
                           "Year: ", Year, "<br>",
                           "Qty: ", format(round(Qty), big.mark = ","), " '000T<br>",
                           "Share: ", Share, "%")]

pce <- ggplot(cereals_yr, aes(x = Year, y = Qty, fill = Commodity,
                               tooltip = tip, data_id = paste(Commodity, Year))) +
  geom_bar_interactive(stat = "identity", position = "stack", width = 0.75, alpha = 0.9) +
  geom_text(aes(label = ifelse(Share >= 8, paste0(Share, "%"), "")),
            position = position_stack(vjust = 0.5), size = 2.8,
            color = "white", fontface = "bold") +
  scale_fill_viridis(discrete = TRUE, option = "D") +
  scale_y_continuous(labels = comma_format()) +
  labs(title   = "Cereal Exports from India by Commodity",
       subtitle = "Rice dominates; wheat exports vary by policy and domestic supply conditions",
       x = "Year", y = "Quantity ('000 Tonnes)",
       caption = "Source: Ministry of Commerce | Analysis: Pawan") +
  my_theme

girafe(ggobj = pce, width_svg = 11, height_svg = 6, options = std_opts)
```

---

## Fertilizer Imports: Composition

```{r fert-imports}
fert <- mqty[HSCode2 == 31]
fert[, Commodity := fcase(
  HSCode4 == "3102", "Nitrogenous (Urea & Others)",
  HSCode4 == "3104", "Potassic (MOP)",
  HSCode4 == "3105", "DAP & NPK Compounds",
  default = "Other Fertilizers"
)]
fert_yr <- fert[, .(Qty = sum(Import, na.rm = TRUE)), by = .(Year, Commodity)]
fert_yr[, Total := sum(Qty), by = Year]
fert_yr[, Share := round(Qty * 100 / Total)]
fert_yr[, tip := paste0("<b>", Commodity, "</b><br>",
                        "Year: ", Year, "<br>",
                        "Qty: ", format(round(Qty), big.mark = ","), " '000T<br>",
                        "Share: ", Share, "%")]

pfe <- ggplot(fert_yr, aes(x = Year, y = Qty, fill = Commodity,
                            tooltip = tip, data_id = paste(Commodity, Year))) +
  geom_bar_interactive(stat = "identity", position = "stack", width = 0.75, alpha = 0.9) +
  geom_text(aes(label = ifelse(Share >= 8, paste0(Share, "%"), "")),
            position = position_stack(vjust = 0.5), size = 2.8,
            color = "white", fontface = "bold") +
  scale_fill_viridis(discrete = TRUE, option = "C") +
  scale_y_continuous(labels = comma_format()) +
  labs(title   = "Fertilizer Imports by Type",
       subtitle = "Nitrogenous (urea-based) fertilizers dominate imports by volume",
       x = "Year", y = "Quantity ('000 Tonnes)",
       caption = "Source: Ministry of Commerce | Analysis: Pawan") +
  my_theme

girafe(ggobj = pfe, width_svg = 11, height_svg = 6, options = std_opts)
```

---

## Petroleum Imports: Composition

```{r petroleum}
petro <- mqty[HSCode2 == 27]
petro[, Commodity := fcase(
  HSCode4 == "2701", "Coal",
  HSCode4 == "2709", "Crude Petroleum Oils",
  HSCode4 == "2711", "Petroleum Gases (LNG/LPG)",
  default = "Other Mineral Fuels"
)]
petro_yr <- petro[, .(Qty = sum(Import, na.rm = TRUE)), by = .(Year, Commodity)]
petro_yr[, Total := sum(Qty), by = Year]
petro_yr[, Share := round(Qty * 100 / Total)]
petro_yr[, tip := paste0("<b>", Commodity, "</b><br>",
                         "Year: ", Year, "<br>",
                         "Qty: ", format(round(Qty), big.mark = ","), " '000T<br>",
                         "Share: ", Share, "%")]

ppe <- ggplot(petro_yr, aes(x = Year, y = Qty, fill = Commodity,
                             tooltip = tip, data_id = paste(Commodity, Year))) +
  geom_bar_interactive(stat = "identity", position = "stack", width = 0.75, alpha = 0.9) +
  geom_text(aes(label = ifelse(Share >= 8, paste0(Share, "%"), "")),
            position = position_stack(vjust = 0.5), size = 2.8,
            color = "white", fontface = "bold") +
  scale_fill_manual(values = c("Crude Petroleum Oils"    = "#e67e22",
                               "Coal"                     = "#2c3e50",
                               "Petroleum Gases (LNG/LPG)" = "#8e44ad",
                               "Other Mineral Fuels"       = "#bdc3c7")) +
  scale_y_continuous(labels = comma_format()) +
  labs(title   = "Petroleum & Mineral Fuel Imports by Type",
       subtitle = "Crude petroleum dominates; LNG imports rising as gas infrastructure expands",
       x = "Year", y = "Quantity ('000 Tonnes)",
       caption = "Source: Ministry of Commerce | Analysis: Pawan") +
  my_theme

girafe(ggobj = ppe, width_svg = 11, height_svg = 6, options = std_opts)
```

---

## Edible Oils & Fats Import Volume

```{r oils-fats}
oils <- mqty[HSCode2 == 15]
oils_yr <- oils[, .(Qty = sum(Import, na.rm = TRUE)), by = Year]
oils_yr[, tip := paste0("Year: <b>", Year, "</b><br>",
                        "Edible oils import: ",
                        format(round(Qty / 1000), big.mark = ","), " '000 Tonnes")]

po <- ggplot(oils_yr, aes(x = Year, y = Qty / 1000, tooltip = tip, data_id = Year, group = 1)) +
  geom_area_interactive(fill = "#f39c12", alpha = 0.3) +
  geom_line_interactive(linewidth = 1.2, color = "#d68910") +
  geom_point_interactive(size = 2.5, color = "#b7770d") +
  scale_y_continuous(labels = comma_format(suffix = " '000T")) +
  labs(title   = "Vegetable Fats & Edible Oils Import Volume",
       subtitle = "Annual import trend — HS Chapter 15",
       x = "Year", y = "Quantity ('000 Tonnes)",
       caption = "Source: Ministry of Commerce | Analysis: Pawan") +
  my_theme

girafe(ggobj = po, width_svg = 10, height_svg = 5, options = std_opts)
```

---

## Key Takeaways

::: {.callout-important title="Summary of Findings"}
1. **India's trade deficit is deep and structural**: Petroleum, fertilizers, and edible oils together account for the bulk of the import bill, and all three are heavily import-dependent by structural necessity.

2. **Cereal exports are policy-sensitive**: The government alternates between encouraging exports (when there is surplus) and banning them (when domestic prices rise) — illustrating the tension between farmer welfare and consumer welfare.

3. **India is the world's largest rice exporter**: Rice contributes the bulk of cereal export earnings, though the recent quality and residue concerns in some markets require attention.

4. **Petroleum imports are the largest single category**: India's energy import dependence (~85% crude oil from overseas) is both an economic vulnerability and a geopolitical consideration.

5. **The export coverage ratio declining** toward 55% suggests imports are growing faster than exports — a medium-term current account concern.

6. **Fertilizer import dependence** reinforces the need to expand domestic production capacity and reduce over-subsidisation, which incentivises excessive fertilizer use and soil degradation.
:::
